---
title: "ALINV – Summary Figures"
date: "`r format(Sys.Date(), '%d %b %Y')`"
format:
  html:
    code-fold: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
  echo: true
  results: hide
output:
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
    toc-expand: true     # <-- always fully open
    code_folding: hide
    highlight: espresso
    theme: journal
    fig_width: 10
    fig_height: 10
---

```{r setup, include=FALSE}
# Knitr settings
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  results = 'hide',  # hides console output
  fig.show = 'hold'  # shows plots only
)

# Functions
source("./functions/_source.R")
source("./functions/0-summary_figures.R")

# Global choice for soil type used by tree-level plots:
soil_type <- "both"        # "both", "inoc-beech", or "inoc-robinia"
```

# Summary of Results for Soil Type: ==`r soil_type`==

# Soil differences

## Soil water content (SWC)
```{r swc, fig.width=10, fig.height=5}
plot_ts_swc(filter_soiltype = soil_type,
            style = "band",
            drought_bars = TRUE,
            save_fig = FALSE, device = "pdf")
```

## Soil respiration (inoc-beech soil type only)
```{r respiration}
plot_ts_resp(filter_soiltype = soil_type,
            style = "band",
            drought_bars = TRUE,
            save_fig = FALSE, device = "pdf")
```

## Soil C/N & isotopes

```{r}
boxplot_soil_cn_isotopes_simple()
```

```{r soil-cn-isotopes}
boxplot_soil_cn_isotopes(save_fig = FALSE, device = "pdf") 
```



# Growing season start

## Phenology
```{r phenology-band}
plot_ts_phenology(style = "band",
                  filter_soiltype = soil_type,
                  save_fig = FALSE, device = "pdf"
                  )
```

## Specific leaf area (SLA)
```{r sla}
boxplot_sla(save_fig = FALSE, device = "pdf")
```

# Growth timeseries

## Height increment
```{r height-inc}
plot_ts_height_inc(filter_soiltype = soil_type,
                   style = "band",
                   drought_bars = TRUE,
                   save_fig = FALSE, device = "pdf")
```

## Diameter increment
```{r diam-inc}
plot_ts_diam_inc(filter_soiltype = soil_type,
                 style = "band",
                 drought_bars = TRUE,
                 save_fig = FALSE, device = "pdf")
```

# Growth until extreme event

## Up to 2025-07-09 (pre-extreme)
```{r growth-pre-jul09}
boxplot_growth_increment_on("2025-07-09", "height",
                            filter_soiltype = soil_type,
                            save_fig = FALSE, device = "pdf") 

boxplot_growth_increment_on("2025-07-09", "diameter",
                            filter_soiltype = soil_type,
                            save_fig = FALSE, device = "pdf") 
```

## Up to 2025-09-29 (including extreme)
```{r growth-incl-sep29}
boxplot_growth_increment_on("2025-09-29", "height",
                            filter_soiltype = soil_type,
                            save_fig = FALSE, device = "pdf") 

boxplot_growth_increment_on("2025-09-29", "diameter",
                            filter_soiltype = soil_type,
                            save_fig = FALSE, device = "pdf") 
```

# Performance during extreme event

## Quantum yield (QY)
```{r qy}
plot_ts_qy(filter_soiltype = soil_type,
           style = "band",
           drought_bars = TRUE,
           save_fig = FALSE, device = "pdf")
```

## Chlorophyll
```{r chl}
plot_ts_chl(filter_soiltype = soil_type,
            style = "band",
            drought_bars = TRUE,
            save_fig = FALSE, device = "pdf")
```

## Tree condition (-)
```{r condition}
plot_ts_condition(filter_soiltype = soil_type,
                  style = "band",
                  drought_bars = TRUE,
                  save_fig = FALSE, device = "pdf")
```

# Senescence

## Senescence (%)
```{r senescence, fig.width=10, fig.height=5}
plot_ts_senescence_percent(filter_soiltype = soil_type,
                           style = "band",
                           drought_bars = FALSE,
                           save_fig = FALSE, device = "pdf") 
```

## Senescence Chlorophyll
```{r, fig.width=10, fig.height=5}
plot_ts_senescence_chl(filter_soiltype = soil_type,
                           style = "band",
                           drought_bars = FALSE,
                           save_fig = FALSE, device = "pdf") 
```

