---
title: 'ALINV – Treatment Effects: Timeseries & Modeling Results'
date: "`r format(Sys.Date(), '%d %b %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
    highlight: espresso
    theme: journal
    fig_width: 9
    fig_height: 6
execute:
  warning: false
  message: false
  echo: true
  results: hide
params:
  scenario_label: "both soils (with soil as treatment)"
  soil_filter: "both"
  include_soil_treatment: true
  analysis_date: !r as.character(Sys.Date())
  output_root: "output"
---

```{r setup, include=FALSE}
# Knitr settings
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  results = 'hide',
  fig.show = 'hold',
  fig.width = 9,
  fig.height = 6
)

# Resolve project root robustly from the notebook render context
source_path_candidates <- c(
  "functions/_source.R",
  "../functions/_source.R",
  "./functions/_source.R"
)
source_path <- source_path_candidates[file.exists(source_path_candidates)][1]
if (is.na(source_path)) {
  stop("Could not locate functions/_source.R for notebook setup.")
}
source(source_path)

source_dir <- dirname(source_path)
proj_root <- normalizePath(file.path(source_dir, ".."), winslash = "/", mustWork = TRUE)

# Function libraries
source(file.path(source_dir, "1-summary-figures.R"))
source(file.path(source_dir, "2-biomass.R"))
source(file.path(source_dir, "3-effect-size-factorial.R"))
source(file.path(source_dir, "4-structural-equation-model.R"))
source(file.path(source_dir, "6-effect-size-model-aggregation.R"))
source(file.path(source_dir, "5-combining-glmm-sem.R"))
source(file.path(source_dir, "7-phenology-transition-models.R"))
source(file.path(source_dir, "8-size-trajectories.R"))

library(patchwork)

# Scenario context
analysis_ctx <- alinv_init_notebook_context(if (exists("params")) params else NULL)
scenario_label <- analysis_ctx$scenario_label
soil_type <- analysis_ctx$soil_filter
include_soil_treatment <- analysis_ctx$include_soil_treatment
analysis_root <- analysis_ctx$analysis_root
analysis_data_root <- analysis_ctx$analysis_data_root
species_vec <- c("fagus", "quercus")

options(alinv.temporal_y_limits = alinv_temporal_effect_y_limits())
```

# Introduction

This notebook presents a comprehensive analysis of the ALINV (Alien Invasion) experiment, which investigates how drought, species mixing, *Robinia* presence, and legacy soil type affect tree physiology and growth. The experiment uses a full-factorial design with two target species — European beech (*Fagus sylvatica*) and sessile oak (*Quercus petraea*) — grown in mesocosm boxes under controlled treatment combinations.

Current render scenario: `r scenario_label`. Soil filtering is set to `r soil_type`, and soil is `r if (include_soil_treatment) "included as an explicit treatment contrast." else "pooled across both soils and omitted as a treatment contrast."`

**Contents:**

1. **Summary** — overall temporal treatment importance, total SEM heatmaps, biomass, and soil conditions.
2. **Specific Results** — response-wise timeseries, temporal GLMMs, and direct SEM panels for each tree trait.

## Methods

### Temporal GLMM (time-varying treatment effects)

For each tree-level response variable, a linear mixed-effects model is fitted across all measurement dates:

$$y_{ijk} = \mu + \alpha_t + (\alpha_t \times \text{Robinia})_{j} + (\alpha_t \times \text{Precipitation})_{j} + (\alpha_t \times \text{Culture})_{j} + (\alpha_t \times \text{Soiltype})_{j} + b_{\text{box}} + b_{\text{tree}} + \varepsilon_{ijk}$$

where $\alpha_t$ is a date fixed effect and treatment effects are estimated per date via date×treatment interactions. Random intercepts are fitted for box ($b_{\text{box}}$) and tree ($b_{\text{tree}}$). Treatment contrasts at each date are extracted as `treatment - baseline`, then oriented for display so positive values indicate beneficial effects and negative values indicate harmful effects.

For this scenario, soil terms are `r if (include_soil_treatment) "included" else "omitted"` in the temporal GLMM.

### Piecewise Structural Equation Model (SEM)

Two-equation piecewise SEM fitted separately per species and trait, using `piecewiseSEM::psem()`:

**SWC sub-model:**
$$\text{SWC} = \beta_0 + \beta_1 \text{DOY} + \beta_2 \text{DOY}^2 + \beta_3 \text{Precipitation} + \beta_4 \text{Robinia} + \beta_5 \text{Soiltype} + \beta_6 \text{Culture} + \beta_7 \text{ExtremeEvent} + b_{\text{box}} + \varepsilon$$

**Response sub-model:**
$$y = \gamma_0 + \gamma_1 \text{DOY} + \gamma_2 \text{DOY}^2 + \gamma_3 \text{SWC} + \gamma_4 \text{Precipitation} + \cdots + b_{\text{box}} + b_{\text{tree}} + \varepsilon$$

Both are `lme4::lmer()` models. Treatment effects are decomposed into:

- **Direct path** ($c'$): treatment → response
- **Indirect path** ($a \times b$): treatment → SWC → response
- **Total effect**: $c' + a \times b$

All numeric predictors are centred and scaled before fitting. For display, response-directed paths are oriented the same way as the temporal GLMM effects and the heatmaps, so positive values indicate beneficial effects and negative values indicate harmful effects.

For this scenario, soil terms are `r if (include_soil_treatment) "included" else "omitted"` in the SEM.

### Biomass GLMM

A single-timepoint linear mixed model for each biomass metric (shoot, root, root:shoot ratio), fitted on z-scored responses to produce standardised coefficients:

$$y_z = \beta_0 + \beta_1 \text{Precipitation} + \beta_2 \text{Culture} + \beta_3 \text{Robinia} + \beta_4 \text{Soiltype} + b_{\text{box}} + \varepsilon$$

Coefficients are in standard deviation units of the response. 95% Wald confidence intervals are used to assess significance.

For this scenario, soil terms are `r if (include_soil_treatment) "included" else "omitted"` in the biomass GLMM.

---

# Summary

```{r notebook-helpers, include=FALSE}
summary_resp_labels <- alinv_response_labels()

summary_biomass_labels <- c(
  shoot_biomass = "Shoot biomass",
  root_biomass = "Root biomass",
  root_shoot_biomass = "Root:shoot biomass"
)

summary_sem_grid <- get_data_var_grid() %>%
  dplyr::filter(.data$type == "tree")

collect_summary_glmm_performance <- function() {
  out <- vector("list", nrow(summary_sem_grid) * length(species_vec))
  idx <- 0L

  for (k in seq_len(nrow(summary_sem_grid))) {
    g <- summary_sem_grid[k, ]

    for (species_name in species_vec) {
      glmm_res <- tryCatch(
        make_effect_figure_generic(
          type = g$type,
          data_name = g$data,
          resp_var = g$resp_var,
          target_species = species_name,
          soil_type = soil_type,
          include_soil_treatment = include_soil_treatment,
          swc_source = "measured",
          force_run = FALSE
        ),
        error = function(e) {
          message("Skipping summary temporal GLMM fit: ", conditionMessage(e))
          NULL
        }
      )

      if (is.null(glmm_res)) next

      idx <- idx + 1L
      out[[idx]] <- extract_temporal_glmm_performance(
        result = glmm_res,
        data_name = g$data,
        resp_var = g$resp_var,
        target_species = species_name,
        soil_type = soil_type,
        include_soil_treatment = include_soil_treatment,
        swc_source = "measured"
      )
    }
  }

  dplyr::bind_rows(out) %>%
    dplyr::mutate(
      response = dplyr::coalesce(unname(summary_resp_labels[.data$resp_var]), .data$resp_var)
    )
}

collect_summary_sem_performance <- function() {
  out <- vector("list", nrow(summary_sem_grid) * length(species_vec))
  idx <- 0L

  for (k in seq_len(nrow(summary_sem_grid))) {
    g <- summary_sem_grid[k, ]

    for (species_name in species_vec) {
      sem_res <- tryCatch(
        run_sem_for_trait(
          type = g$type,
          data_name = g$data,
          resp_var = g$resp_var,
          species = species_name,
          soil_type = soil_type,
          include_soil_treatment = include_soil_treatment,
          include_interaction = FALSE,
          phase_window = "all",
          scale_all_numeric = TRUE,
          swc_source = "measured",
          force_run = FALSE
        ),
        error = function(e) {
          message("Skipping summary SEM performance fit: ", conditionMessage(e))
          NULL
        }
      )

      if (is.null(sem_res)) next

      idx <- idx + 1L
      out[[idx]] <- extract_sem_model_performance(
        result = sem_res,
        data_name = g$data,
        resp_var = g$resp_var,
        species = species_name,
        soil_type = soil_type,
        include_soil_treatment = include_soil_treatment,
        swc_source = "measured"
      )
    }
  }

  dplyr::bind_rows(out) %>%
    dplyr::mutate(
      response = dplyr::coalesce(unname(summary_resp_labels[.data$resp_var]), .data$resp_var)
    )
}

collect_summary_sem_matrix_data <- function(species_name) {
  out <- vector("list", nrow(summary_sem_grid))

  for (k in seq_len(nrow(summary_sem_grid))) {
    g <- summary_sem_grid[k, ]

    sem_res <- tryCatch(
      run_sem_for_trait(
        type = g$type,
        data_name = g$data,
        resp_var = g$resp_var,
        species = species_name,
        soil_type = soil_type,
        include_soil_treatment = include_soil_treatment,
        include_interaction = FALSE,
        phase_window = "all",
        scale_all_numeric = TRUE,
        swc_source = "measured",
        force_run = FALSE
      ),
      error = function(e) {
        message("Skipping summary SEM fit: ", conditionMessage(e))
        NULL
      }
    )

    if (is.null(sem_res)) next

    matrix_df <- sem_res$matrix_data %||% tibble::tibble()
    if (!nrow(matrix_df) && !is.null(sem_res$effects) && nrow(sem_res$effects)) {
      matrix_df <- build_sem_matrix_data(
        effects_main = sem_res$effects,
        effects_int = sem_res$effects_int,
        resp_var = g$resp_var,
        species = species_name,
        include_interaction = FALSE,
        swc_source = "measured",
        phase = "all"
      )
    }

    if (!nrow(matrix_df)) next

    out[[k]] <- matrix_df
  }

  dplyr::bind_rows(out)
}

plot_summary_total_sem_heatmaps <- function() {
  sem_all_species <- purrr::map_dfr(species_vec, collect_summary_sem_matrix_data)

  if (!nrow(sem_all_species)) {
    return(
      ggplot() +
        theme_void() +
        ggtitle("No total SEM heatmaps available")
    )
  }

  shared_limit <- compute_shared_sem_heatmap_limit(sem_all_species)

  plot_list <- purrr::map(
    species_vec,
    function(species_name) {
      panel_df <- prepare_sem_heatmap_panel(
        matrix_df = sem_all_species %>%
          dplyr::filter(.data$species == species_name, .data$effect_class == "main"),
        path_type = "total",
        resp_labels = summary_resp_labels,
        treat_labels = sem_heatmap_treatment_labels()
      )

      plot_sem_heatmap_panel(
        panel_df = panel_df,
        limit = shared_limit,
        title = paste0(stringr::str_to_title(species_name), ": Total SEM effects"),
        annotate_values = TRUE,
        value_digits = 2
      )
    }
  )

  patchwork::wrap_plots(plot_list, ncol = 1)
}

plot_summary_volume_sem_combo <- function(growth_resp_var = "volume_inc_phase_abs") {
  sem_all_species <- purrr::map_dfr(species_vec, collect_summary_sem_matrix_data)

  if (!nrow(sem_all_species)) {
    return(
      ggplot() +
        theme_void() +
        ggtitle("No SEM matrix data available for combined volume summary")
    )
  }

  plot_volume_sem_summary_figure(
    matrix_df = sem_all_species,
    growth_resp_var = growth_resp_var,
    soil_type = soil_type,
    include_soil_treatment = include_soil_treatment,
    species_vec = species_vec
  )
}

df_biomass <- wrangle_tree_biomass(file.path(proj_root, "data", "raw", "data_20260309.xlsx"))
res_fag <- run_biomass_glmm_species(
  df_biomass,
  species_keep = "fagus",
  soil_type = soil_type,
  include_soil_treatment = include_soil_treatment
)
res_que <- run_biomass_glmm_species(
  df_biomass,
  species_keep = "quercus",
  soil_type = soil_type,
  include_soil_treatment = include_soil_treatment
)

summary_glmm_performance <- dplyr::bind_rows(
  collect_summary_glmm_performance(),
  dplyr::bind_rows(res_fag$performance, res_que$performance) %>%
    dplyr::transmute(
      model_scope = "Biomass GLMM",
      data_name = "biomass",
      resp_var = .data$metric,
      response = dplyr::coalesce(unname(summary_biomass_labels[.data$metric]), .data$metric),
      species = .data$species,
      soil_filter = soil_type,
      include_soil_treatment = isTRUE(include_soil_treatment),
      swc_source = NA_character_,
      n_obs = .data$n_obs,
      n_boxes = NA_integer_,
      n_trees = NA_integer_,
      n_dates = NA_integer_,
      aic = .data$AIC,
      bic = .data$BIC,
      logLik = NA_real_,
      sigma = NA_real_,
      r2_marginal = .data$R2_marginal,
      r2_conditional = .data$R2_conditional
    )
) %>%
  dplyr::arrange(.data$model_scope, .data$species, .data$response)

summary_sem_performance <- collect_summary_sem_performance() %>%
  dplyr::arrange(.data$species, .data$response)

```

## Temporal Effects

Aggregated temporal GLMM effect sizes across all tree response variables, grouped by experimental phase (until June / July–August / September+).

```{r treatment-effects-overview, fig.width=6, fig.height=12}
# Collect all temporal GLMM effects across variables and species
all_eff <- collect_all_effects(
  var_grid    = get_data_var_grid() %>% dplyr::filter(type == "tree"),
  species_vec = c("fagus", "quercus"),
  soil_type   = soil_type,
  include_soil_treatment = include_soil_treatment
)

if (nrow(all_eff) > 0) {
  all_eff_imp <- all_eff %>% 
    add_importance_metric(use_stat = FALSE, scale_within = TRUE)
  all_eff_imp_phase <- add_phase_to_effects(all_eff_imp)
  
  # Overall effects across all species
  p_all <- plot_treatment_importance_phases(
    all_eff_imp_phase, 
    stat = "mean_se"
  ) + coord_cartesian(ylim = c(0, 3)) + ggtitle("All species")
  
  # Fagus only
  p_fag <- plot_treatment_importance_phases(
    all_eff_imp_phase %>% dplyr::filter(species == "fagus"),
    stat = "mean_se"
  ) + coord_cartesian(ylim = c(0, 3)) + ggtitle("Fagus")
  
  # Quercus only
  p_que <- plot_treatment_importance_phases(
    all_eff_imp_phase %>% dplyr::filter(species == "quercus"),
    stat = "mean_se"
  ) + coord_cartesian(ylim = c(0, 3)) + ggtitle("Quercus")
  
  p_combined <- (p_all / p_fag / p_que) +
    patchwork::plot_layout(guides = "collect") +
    patchwork::plot_annotation(tag_levels = "A") &
    ggplot2::theme(legend.position = "bottom",
                   plot.tag = ggplot2::element_text(face = "bold"))
  
  print(p_combined)
} else {
  message("No effects data collected. This may happen on first run.")
}
```

---

## Total SEM Effects

These heatmaps summarise the total SEM effect of each treatment on each response variable, using the shared color scale from the full-data SEM fits in the current scenario. Positive values indicate beneficial effects and negative values indicate harmful effects. For senescence, the inferential panels labeled `Senescence` are fitted on `Remaining green (%) = 100 - Senescence (%)`, so negative values indicate earlier or faster senescence progression.

```{r total-sem-effects-summary, fig.width=12, fig.height=8}
print(plot_summary_total_sem_heatmaps())
```

---

## Phase-Wise Allometric-Volume Growth and Total SEM Effects

This combined overview links the phase-based absolute increment of `Volume` to a reduced total-effect SEM heatmap. `Volume` is estimated from the Annighöfer et al. (2016) Table 4 `RCD^2H` allometry and shown here only as the phase-based absolute increment among the growth metrics.

```{r volume-sem-summary-combo, fig.width=16, fig.height=9}
volume_sem_combo_plot <- plot_summary_volume_sem_combo(growth_resp_var = "volume_inc_phase_abs")
print(volume_sem_combo_plot)

save_growth_plot_png(
  plot_obj = volume_sem_combo_plot,
  file_path = file.path(
    alinv_data_path("summary_figures", create_dir = TRUE),
    "volume-phase-sem-summary.png"
  ),
  width = 16,
  height = 9
)
```

---

## Phase-Wise Relative Volume Growth and Total SEM Effects

This combined overview links the phase-based relative `Volume` increment timeseries to a reduced total-effect SEM heatmap. `Volume` is estimated from the Annighöfer et al. (2016) Table 4 `RCD^2H` allometry and shown here only as the phase-based relative increment among the growth metrics.

```{r volume-sem-summary-combo-relative, fig.width=16, fig.height=9}
volume_sem_combo_rel_plot <- plot_summary_volume_sem_combo(growth_resp_var = "volume_inc_phase_rel")
print(volume_sem_combo_rel_plot)

save_growth_plot_png(
  plot_obj = volume_sem_combo_rel_plot,
  file_path = file.path(
    alinv_data_path("summary_figures", create_dir = TRUE),
    "volume-phase-rel-sem-summary.png"
  ),
  width = 16,
  height = 9
)
```

---

## Model Performance

The tables below summarise the fitted-model performance metrics across the notebook. Temporal and biomass GLMMs report AIC/BIC and marginal/conditional $R^2$ when available. SEMs report the SWC and response submodel AIC/BIC values plus SEM $R^2$ values when `piecewiseSEM` is available in the current R session.

### GLMM Performance

```{r glmm-performance-summary, results='asis'}
knitr::kable(
  summary_glmm_performance %>%
    dplyr::select(model_scope, species, response, n_obs, n_dates, aic, bic, r2_marginal, r2_conditional) %>%
    dplyr::mutate(
      dplyr::across(
        c(aic, bic, r2_marginal, r2_conditional),
        ~ round(.x, 3)
      )
    ),
  caption = "GLMM performance summary"
)
```

### SEM Performance

```{r sem-performance-summary, results='asis'}
knitr::kable(
  summary_sem_performance %>%
    dplyr::select(
      species, response, n_obs,
      aic_swc, bic_swc, aic_response, bic_response,
      r2_swc_marginal, r2_swc_conditional,
      r2_response_marginal, r2_response_conditional,
      piecewise_r2_available
    ) %>%
    dplyr::mutate(
      dplyr::across(
        c(aic_swc, bic_swc, aic_response, bic_response,
          r2_swc_marginal, r2_swc_conditional,
          r2_response_marginal, r2_response_conditional),
        ~ round(.x, 3)
      )
    ),
  caption = "SEM performance summary"
)
```

---

## Biomass

```{r biomass-data, include=FALSE}
# Biomass data and fitted biomass GLMMs are initialized in `notebook-helpers`
# so the summary performance table and the biomass section use the same results.
```

### Fagus Boxplots

```{r biomass-boxplot-fagus, fig.width=9, fig.height=10}
print(plot_tree_biomass_treatments(df_biomass, species_keep = "fagus"))
```

### Quercus Boxplots

```{r biomass-boxplot-quercus, fig.width=9, fig.height=10}
print(plot_tree_biomass_treatments(df_biomass, species_keep = "quercus"))
```

### Biomass GLMM

A one-off GLMM (no temporal factor) quantifies the influence of precipitation, culture, robinia presence, and soil type on root, shoot, and root:shoot biomass.

### Fagus

```{r biomass-glmm-fagus, fig.width=8, fig.height=6}
print(res_fag$plot)
```

```{r biomass-glmm-fagus-perf, results='asis'}
knitr::kable(res_fag$performance %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))),
  caption = "Fagus biomass GLMM performance")
```

### Quercus

```{r biomass-glmm-quercus, fig.width=8, fig.height=6}
print(res_que$plot)
```

```{r biomass-glmm-quercus-perf, results='asis'}
knitr::kable(res_que$performance %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))),
  caption = "Quercus biomass GLMM performance")
```

---

## Soil Water Availability

### Soil Water Content

```{r swc, fig.width=12, fig.height=6}
plot_ts_swc(filter_soiltype = soil_type,
            style = "band",
            drought_bars = TRUE,
            save_fig = FALSE, device = "pdf")
```

### Soil Water Potential

#### Raw Data + LOESS Smoothed
```{r water-potential-both, fig.width=10, fig.height=6}
plot_ts_soil_mp(
  filter_soiltype = soil_type,
  show="both",
  save_fig = FALSE,
  device = "pdf"
)
```

#### Raw Data
```{r water-potential-raw, fig.width=10, fig.height=6}
plot_ts_soil_mp(
  filter_soiltype = soil_type,
  show="raw",
  save_fig = FALSE,
  device = "pdf"
)
```


### Soil Respiration

```{r respiration, fig.width=12, fig.height=6}
plot_ts_resp(filter_soiltype = soil_type,
            style = "band",
            drought_bars = TRUE,
            save_fig = FALSE, device = "pdf")
```


### Soil C/N and Isotopes

```{r isotopes, fig.width=12, fig.height=10}
boxplot_soil_cn_isotopes(save_fig = FALSE, device = "pdf")
```

---

# Specific Results

## Chlorophyll

### Timeseries

```{r chl-ts, fig.width=9.5, fig.height=5.8}
plot_ts_chl(filter_soiltype = soil_type,
            style = "band",
            drought_bars = TRUE,
            save_fig = FALSE, device = "pdf")
```

### Fagus

#### Temporal Effects

```{r chl-fagus-effect-size, fig.width=9.5, fig.height=5.8}
res_chl_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "chlorophyll",
  resp_var = "chl",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_chl_fagus$temporal$plot)
```

#### SEM

```{r chl-fagus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_chl_fagus$sem$plot)
```

### Quercus

#### Temporal Effects

```{r chl-quercus-effect-size, fig.width=9.5, fig.height=5.8}
res_chl_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "chlorophyll",
  resp_var = "chl",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_chl_quercus$temporal$plot)
```

#### SEM

```{r chl-quercus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_chl_quercus$sem$plot)
```

## Condition

### Timeseries

```{r cond-ts, fig.width=9.5, fig.height=5.8}
plot_ts_condition(filter_soiltype = soil_type,
                  style = "band",
                  drought_bars = TRUE,
                  save_fig = FALSE, device = "pdf")
```

### Fagus

#### Temporal Effects

```{r cond-fagus-effect-size, fig.width=9.5, fig.height=5.8}
res_cond_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "condition",
  resp_var = "condition",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_cond_fagus$temporal$plot)
```

#### SEM

```{r cond-fagus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_cond_fagus$sem$plot)
```

### Quercus

#### Temporal Effects

```{r cond-quercus-effect-size, fig.width=9.5, fig.height=5.8}
res_cond_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "condition",
  resp_var = "condition",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_cond_quercus$temporal$plot)
```

#### SEM

```{r cond-quercus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_cond_quercus$sem$plot)
```

## Growth

The growth chapter covers `Height`, `Diameter`, and `Volume`, each split into absolute size, absolute increment from the first measurement, relative increment from the first measurement, absolute increment within phase, and relative increment within phase. `Volume` is estimated from the Annighöfer et al. (2016) Table 4 `RCD^2H` allometry rather than from a cylindrical stem formula. For the phase-baseline metrics, phase 1 is scaled to the first measurement and later phases are scaled to the last measurement of the previous phase. The artificial zero-valued restart points are used only to clarify the within-phase timeseries and are never used for temporal GLMMs, SEMs, caches, or heatmap exports.

```{r growth-figure-manifest, include=FALSE}
growth_specs_main <- growth_metric_catalog()
growth_figures_main <- build_growth_figure_manifest(
  soil_type = soil_type,
  include_soil_treatment = include_soil_treatment
)

readr::write_csv(
  growth_figures_main,
  file.path(alinv_data_path("size_trajectories", create_dir = TRUE), "size-trajectories-figure-manifest.csv")
)
```

```{r growth-response-loop, results='asis', echo=FALSE}
for (metric_name in unique(growth_specs_main$metric_group)) {
  cat("### ", metric_name, "\n\n", sep = "")

  specs_metric <- growth_specs_main %>%
    dplyr::filter(.data$metric_group == metric_name)

  for (i in seq_len(nrow(specs_metric))) {
    spec_i <- specs_metric[i, ]

    cat("#### ", spec_i$variant_label[[1]], "\n\n", sep = "")
    cat("##### Timeseries\n\n")
    cat(
      growth_markdown_image(
        get_growth_figure_path(
          manifest = growth_figures_main,
          figure_type = "timeseries",
          resp_var = spec_i$resp_var[[1]]
        )
      )
    )

    for (species_i in species_vec) {
      cat("##### ", stringr::str_to_title(species_i), "\n\n", sep = "")
      cat("###### Temporal GLMM\n\n")
      cat(
        growth_markdown_image(
          get_growth_figure_path(
            manifest = growth_figures_main,
            figure_type = "temporal",
            resp_var = spec_i$resp_var[[1]],
            species = species_i
          )
        )
      )
      cat("###### SEM\n\n")
      cat(
        growth_markdown_image(
          get_growth_figure_path(
            manifest = growth_figures_main,
            figure_type = "sem",
            resp_var = spec_i$resp_var[[1]],
            species = species_i
          )
        )
      )
    }
  }
}
```

## Phenology

### Timeseries

```{r phenology-ts, fig.width=9.5, fig.height=5.8}
plot_ts_phenology(style = "band",
                  filter_soiltype = soil_type,
                  save_fig = FALSE, device = "pdf")
```

The phenology timeseries above is descriptive: it shows the mean day of year (DOY) at which each treatment group reaches stages 1-4. The uncertainty bars show `mean ± SE` for the transition timing at each stage.

To answer the more direct biological question, “which treatment makes trees reach a given stage earlier or later?”, we additionally fit transition-timing models using the DOY of reaching stages 2, 3, and 4 as the response. For each species and stage separately, this is a linear mixed model of the form `DOY ~ precipitation + robinia + culture (+ soiltype, when included) + (1 | boxlabel)`. The “Average” column comes from a pooled mixed model across stages 2-4, `DOY ~ stage + precipitation + robinia + culture (+ soiltype) + (1 | boxlabel) + (1 | tree_id)`, so that the mean timing difference among stages is accounted for while repeated stage transitions from the same tree are kept together.

The plotted points are therefore estimated treatment shifts in transition date, in units of days, with 95% Wald confidence intervals from the fixed-effect coefficients. Lower DOY means earlier stage attainment, so negative values indicate that the named treatment advances the stage transition and positive values indicate that it delays it.

Scenario-local caches and CSV exports for these transition-timing models are written under `r alinv_project_relative_path(file.path(analysis_data_root, "model-phenology-transition"))`.

```{r phenology-transition-summary, fig.width=13.5, fig.height=6.8}
phenology_transition_results <- purrr::map(
  c("fagus", "quercus"),
  ~ run_phenology_transition_analysis(
    species_keep = .x,
    soil_type = soil_type,
    include_soil_treatment = include_soil_treatment,
    force_run = FALSE
  )
)

phenology_transition_effects <- dplyr::bind_rows(
  purrr::map(phenology_transition_results, "effects")
)

print(plot_phenology_transition_effects(phenology_transition_effects))
```

The temporal GLMM and SEM below keep the original progression-model view, where the response is the current phenological stage observed at a given date. Higher predicted stage values therefore mean more advanced leaf development at that point in time.

### Fagus

```{r phenology-fagus-panels, fig.width=9.5, fig.height=5.8}
res_phenology_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "phenology",
  resp_var = "stage",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_phenology_fagus$temporal$plot)
print(res_phenology_fagus$sem$plot)
```

### Quercus

```{r phenology-quercus-panels, fig.width=9.5, fig.height=5.8}
res_phenology_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "phenology",
  resp_var = "stage",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_phenology_quercus$temporal$plot)
print(res_phenology_quercus$sem$plot)
```

## Quantum Yield

### Timeseries

```{r qy-ts, fig.width=9.5, fig.height=5.8}
plot_ts_qy(filter_soiltype = soil_type,
           style = "band",
           drought_bars = TRUE,
           save_fig = FALSE, device = "pdf")
```

### Fagus

```{r qy-fagus-panels, fig.width=9.5, fig.height=5.8}
res_qy_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "quantum_yield",
  resp_var = "qy",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_qy_fagus$temporal$plot)
print(res_qy_fagus$sem$plot)
```

### Quercus

```{r qy-quercus-panels, fig.width=9.5, fig.height=5.8}
res_qy_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "quantum_yield",
  resp_var = "qy",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_qy_quercus$temporal$plot)
print(res_qy_quercus$sem$plot)
```

## Senescence

### Percent Senesced

#### Timeseries

```{r senescence-pct-ts, fig.width=9.5, fig.height=5.8}
plot_ts_senescence_percent(filter_soiltype = soil_type,
                           style = "band",
                           drought_bars = FALSE,
                           save_fig = FALSE, device = "pdf")
```

#### Fagus

The raw timeseries above shows measured `Senescence (%)`. The temporal GLMM and SEM panels below are labeled `Senescence`, but are fitted on `Remaining green (%) = 100 - Senescence (%)`, so negative effects indicate earlier or faster senescence progression. This is not a true senescence-onset timing model.

```{r senescence-pct-fagus-panels, fig.width=9.5, fig.height=5.8}
res_sen_pct_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "remaining_green",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_sen_pct_fagus$temporal$plot)
print(res_sen_pct_fagus$sem$plot)
```

#### Quercus

```{r senescence-pct-quercus-panels, fig.width=9.5, fig.height=5.8}
res_sen_pct_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "remaining_green",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_sen_pct_quercus$temporal$plot)
print(res_sen_pct_quercus$sem$plot)
```

### Chlorophyll

#### Timeseries

```{r senescence-chl-ts, fig.width=9.5, fig.height=5.8}
plot_ts_senescence_chl(filter_soiltype = soil_type,
                       style = "band",
                       drought_bars = FALSE,
                       save_fig = FALSE, device = "pdf")
```

#### Fagus

```{r senescence-chl-fagus-panels, fig.width=9.5, fig.height=5.8}
res_sen_chl_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "chlavg",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_sen_chl_fagus$temporal$plot)
print(res_sen_chl_fagus$sem$plot)
```

#### Quercus

```{r senescence-chl-quercus-panels, fig.width=9.5, fig.height=5.8}
res_sen_chl_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "chlavg",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_sen_chl_quercus$temporal$plot)
print(res_sen_chl_quercus$sem$plot)
```
# Session Info

```{r session-info}
sessionInfo()
```
