---
title: 'ALINV – Treatment Effects: Timeseries & Modeling Results'
date: "`r format(Sys.Date(), '%d %b %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    highlight: espresso
    theme: journal
    fig_width: 9
    fig_height: 6
execute:
  warning: false
  message: false
  echo: true
  results: hide
---

```{r setup, include=FALSE}
# Knitr settings
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  results = 'hide',
  fig.show = 'hold',
  fig.width = 9,
  fig.height = 6
)

# Resolve project root from notebook location (robust to knit working directory)
input_file <- knitr::current_input()
input_dir <- dirname(normalizePath(input_file, winslash = "/", mustWork = TRUE))
proj_root <- normalizePath(file.path(input_dir, ".."), winslash = "/", mustWork = TRUE)

# Function libraries
source(file.path(proj_root, "functions", "_source.R"))
source(file.path(proj_root, "functions", "1-summary-figures.R"))
source(file.path(proj_root, "functions", "2-biomass.R"))
source(file.path(proj_root, "functions", "3-effect-size-factorial.R"))
source(file.path(proj_root, "functions", "4-structural-equation-model.R"))
source(file.path(proj_root, "functions", "6-effect-size-model-aggregation.R"))
source(file.path(proj_root, "functions", "5-combining-glmm-sem.R"))

library(patchwork)

# Global choices
soil_type <- "both"        # "both", "inoc-beech", or "inoc-robinia"
species_vec <- c("fagus", "quercus")

# Global temporal effect y-axis limits (shared across all temporal GLMM figures)
tmp_eff_for_ylim <- tryCatch(
  collect_all_effects(
    var_grid    = get_data_var_grid() %>% dplyr::filter(type == "tree"),
    species_vec = species_vec,
    soil_type   = soil_type
  ),
  error = function(e) tibble::tibble()
)

if (nrow(tmp_eff_for_ylim) > 0) {
  options(alinv.temporal_y_limits = compute_global_effect_y_limits(tmp_eff_for_ylim))
} else {
  options(alinv.temporal_y_limits = NULL)
}
```

# Introduction

This notebook presents a comprehensive analysis of the ALINV (Alien Invasion) experiment, which investigates how drought, species mixing, *Robinia* presence, and legacy soil type affect tree physiology and growth. The experiment uses a full-factorial design with two target species — European beech (*Fagus sylvatica*) and sessile oak (*Quercus petraea*) — grown in mesocosm boxes under controlled treatment combinations.

**Contents:**

1. **Treatment Effects Summary** — aggregated temporal GLMM effect sizes across all response variables, grouped by experimental phase.
2. **Biomass** — end-of-experiment destructive harvest data (shoot, root, root:shoot ratio), visualised as treatment boxplots and quantified with a one-off GLMM.
3. **Soil Conditions** — time series of soil water content (SWC), soil water potential (SWP), soil respiration, and C/N isotopes.
4. **Tree Variables** — for each trait (chlorophyll, condition, height growth, diameter growth, phenology, quantum yield, senescence), the notebook shows: (i) a treatment-level time series, (ii) temporal effect sizes from date-specific GLMMs, and (iii) piecewise structural equation models (SEM) decomposing direct and SWC-mediated indirect effects.

## Methods

### Temporal GLMM (time-varying treatment effects)

For each tree-level response variable, a linear mixed-effects model is fitted across all measurement dates:

$$y_{ijk} = \mu + \alpha_t + (\alpha_t \times \text{Robinia})_{j} + (\alpha_t \times \text{Precipitation})_{j} + (\alpha_t \times \text{Culture})_{j} + (\alpha_t \times \text{Soiltype})_{j} + b_{\text{box}} + b_{\text{tree}} + \varepsilon_{ijk}$$

where $\alpha_t$ is a date fixed effect and treatment effects are estimated per date via date×treatment interactions. Random intercepts are fitted for box ($b_{\text{box}}$) and tree ($b_{\text{tree}}$). Treatment contrasts at each date are extracted using estimated marginal means (`emmeans`, reverse pairwise).

### Piecewise Structural Equation Model (SEM)

Two-equation piecewise SEM fitted separately per species and trait, using `piecewiseSEM::psem()`:

**SWC sub-model:**
$$\text{SWC} = \beta_0 + \beta_1 \text{DOY} + \beta_2 \text{DOY}^2 + \beta_3 \text{Precipitation} + \beta_4 \text{Robinia} + \beta_5 \text{Soiltype} + \beta_6 \text{Culture} + \beta_7 \text{ExtremeEvent} + b_{\text{box}} + \varepsilon$$

**Response sub-model:**
$$y = \gamma_0 + \gamma_1 \text{DOY} + \gamma_2 \text{DOY}^2 + \gamma_3 \text{SWC} + \gamma_4 \text{Precipitation} + \cdots + b_{\text{box}} + b_{\text{tree}} + \varepsilon$$

Both are `lme4::lmer()` models. Treatment effects are decomposed into:

- **Direct path** ($c'$): treatment → response
- **Indirect path** ($a \times b$): treatment → SWC → response
- **Total effect**: $c' + a \times b$

All numeric predictors are centred and scaled before fitting.

### Biomass GLMM

A single-timepoint linear mixed model for each biomass metric (shoot, root, root:shoot ratio), fitted on z-scored responses to produce standardised coefficients:

$$y_z = \beta_0 + \beta_1 \text{Precipitation} + \beta_2 \text{Culture} + \beta_3 \text{Robinia} + \beta_4 \text{Soiltype} + b_{\text{box}} + \varepsilon$$

Coefficients are in standard deviation units of the response. 95% Wald confidence intervals are used to assess significance.

---

# Overview: Treatment Effects Summary

Aggregated temporal GLMM effect sizes across all tree response variables, grouped by experimental phase (until June / July–August / September+).

```{r treatment-effects-overview, fig.width=6, fig.height=12}
# Collect all temporal GLMM effects across variables and species
all_eff <- collect_all_effects(
  var_grid    = get_data_var_grid() %>% dplyr::filter(type == "tree"),
  species_vec = c("fagus", "quercus"),
  soil_type   = "both"
)

if (nrow(all_eff) > 0) {
  all_eff_imp <- all_eff %>% 
    add_importance_metric(use_stat = FALSE, scale_within = TRUE)
  all_eff_imp_phase <- add_phase_to_effects(all_eff_imp)
  
  # Overall effects across all species
  p_all <- plot_treatment_importance_phases(
    all_eff_imp_phase, 
    stat = "mean_se"
  ) + coord_cartesian(ylim = c(0, 3)) + ggtitle("All species")
  
  # Fagus only
  p_fag <- plot_treatment_importance_phases(
    all_eff_imp_phase %>% dplyr::filter(species == "fagus"),
    stat = "mean_se"
  ) + coord_cartesian(ylim = c(0, 3)) + ggtitle("Fagus")
  
  # Quercus only
  p_que <- plot_treatment_importance_phases(
    all_eff_imp_phase %>% dplyr::filter(species == "quercus"),
    stat = "mean_se"
  ) + coord_cartesian(ylim = c(0, 3)) + ggtitle("Quercus")
  
  p_combined <- (p_all / p_fag / p_que) +
    patchwork::plot_layout(guides = "collect") +
    patchwork::plot_annotation(tag_levels = "A") &
    ggplot2::theme(legend.position = "bottom",
                   plot.tag = ggplot2::element_text(face = "bold"))
  
  print(p_combined)
} else {
  message("No effects data collected. This may happen on first run.")
}
```

---

# Biomass: Treatment Comparison

```{r biomass-data}
df_biomass <- wrangle_tree_biomass(file.path(proj_root, "data", "raw", "data_20260309.xlsx"))
```

## Fagus – Boxplots

```{r biomass-boxplot-fagus, fig.width=9, fig.height=10}
print(plot_tree_biomass_treatments(df_biomass, species_keep = "fagus"))
```

## Quercus – Boxplots

```{r biomass-boxplot-quercus, fig.width=9, fig.height=10}
print(plot_tree_biomass_treatments(df_biomass, species_keep = "quercus"))
```

## GLMM: Treatment Effects on Biomass

A one-off GLMM (no temporal factor) quantifies the influence of precipitation, culture, robinia presence, and soil type on root, shoot, and root:shoot biomass.

### Fagus

```{r biomass-glmm-fagus, fig.width=8, fig.height=6}
res_fag <- run_biomass_glmm_species(df_biomass, species_keep = "fagus")
print(res_fag$plot)
```

```{r biomass-glmm-fagus-perf, results='asis'}
knitr::kable(res_fag$performance %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))),
  caption = "Fagus biomass GLMM performance")
```

### Quercus

```{r biomass-glmm-quercus, fig.width=8, fig.height=6}
res_que <- run_biomass_glmm_species(df_biomass, species_keep = "quercus")
print(res_que$plot)
```

```{r biomass-glmm-quercus-perf, results='asis'}
knitr::kable(res_que$performance %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))),
  caption = "Quercus biomass GLMM performance")
```

---

# Context: Soil Conditions

## Soil Water Content (SWC)

```{r swc, fig.width=12, fig.height=6}
plot_ts_swc(filter_soiltype = soil_type,
            style = "band",
            drought_bars = TRUE,
            save_fig = FALSE, device = "pdf")
```

## Soil Water Potential (Matric Potential)

### Raw Data + LOESS Smoothed
```{r water-potential-both, fig.width=10, fig.height=6}
plot_ts_soil_mp(
  filter_soiltype = soil_type,
  show="both",
  save_fig = FALSE,
  device = "pdf"
)
```

### Raw Data
```{r water-potential-raw, fig.width=10, fig.height=6}
plot_ts_soil_mp(
  filter_soiltype = soil_type,
  show="raw",
  save_fig = FALSE,
  device = "pdf"
)
```


## Soil Respiration

```{r respiration, fig.width=12, fig.height=6}
plot_ts_resp(filter_soiltype = soil_type,
            style = "band",
            drought_bars = TRUE,
            save_fig = FALSE, device = "pdf")
```


## Soil C/N & Isotopes

```{r isotopes, fig.width=12, fig.height=10}
boxplot_soil_cn_isotopes(save_fig = FALSE, device = "pdf")
```

---

# Tree Variables: Timeseries, Temporal Models & SEM

## Chlorophyll Content (Chl)

### Timeseries (Beech and Oak)

```{r chl-ts, fig.width=9.5, fig.height=5.8}
plot_ts_chl(filter_soiltype = soil_type,
            style = "band",
            drought_bars = TRUE,
            save_fig = FALSE, device = "pdf")
```

### Analysis Beech (Fagus)

#### Effect Size Over Time

```{r chl-fagus-effect-size, fig.width=9.5, fig.height=5.8}
res_chl_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "chlorophyll",
  resp_var = "chl",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_chl_fagus$temporal$plot)
```

#### SEM Without Interaction

```{r chl-fagus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_chl_fagus$sem$plot)
```

#### SEM With Interaction

```{r chl-fagus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_chl_fagus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "chlorophyll",
  resp_var = "chl",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_chl_fagus_int$sem$plot)
```

### Analysis Oak (Quercus)

#### Effect Size Over Time

```{r chl-quercus-effect-size, fig.width=9.5, fig.height=5.8}
res_chl_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "chlorophyll",
  resp_var = "chl",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_chl_quercus$temporal$plot)
```

#### SEM Without Interaction

```{r chl-quercus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_chl_quercus$sem$plot)
```

#### SEM With Interaction

```{r chl-quercus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_chl_quercus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "chlorophyll",
  resp_var = "chl",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_chl_quercus_int$sem$plot)
```

---

## Tree Condition

### Timeseries (Beech and Oak)

```{r cond-ts, fig.width=9.5, fig.height=5.8}
plot_ts_condition(filter_soiltype = soil_type,
                  style = "band",
                  drought_bars = TRUE,
                  save_fig = FALSE, device = "pdf")
```

### Analysis Beech (Fagus)

#### Effect Size Over Time

```{r cond-fagus-effect-size, fig.width=9.5, fig.height=5.8}
res_cond_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "condition",
  resp_var = "condition",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_cond_fagus$temporal$plot)
```

#### SEM Without Interaction

```{r cond-fagus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_cond_fagus$sem$plot)
```

#### SEM With Interaction

```{r cond-fagus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_cond_fagus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "condition",
  resp_var = "condition",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_cond_fagus_int$sem$plot)
```

### Analysis Oak (Quercus)

#### Effect Size Over Time

```{r cond-quercus-effect-size, fig.width=9.5, fig.height=5.8}
res_cond_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "condition",
  resp_var = "condition",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_cond_quercus$temporal$plot)
```

#### SEM Without Interaction

```{r cond-quercus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_cond_quercus$sem$plot)
```

#### SEM With Interaction

```{r cond-quercus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_cond_quercus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "condition",
  resp_var = "condition",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_cond_quercus_int$sem$plot)
```

---

## Growth: Height

### Timeseries (Beech and Oak)

```{r height-ts, fig.width=9.5, fig.height=5.8}
plot_ts_height_inc(filter_soiltype = soil_type,
                   style = "band",
                   drought_bars = TRUE,
                   save_fig = FALSE, device = "pdf")
```

### Analysis Beech (Fagus)

#### Effect Size Over Time

```{r height-fagus-effect-size, fig.width=9.5, fig.height=5.8}
res_height_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "growth",
  resp_var = "height",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_height_fagus$temporal$plot)
```

#### SEM Without Interaction

```{r height-fagus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_height_fagus$sem$plot)
```

#### SEM With Interaction

```{r height-fagus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_height_fagus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "growth",
  resp_var = "height",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_height_fagus_int$sem$plot)
```

### Analysis Oak (Quercus)

#### Effect Size Over Time

```{r height-quercus-effect-size, fig.width=9.5, fig.height=5.8}
res_height_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "growth",
  resp_var = "height",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_height_quercus$temporal$plot)
```

#### SEM Without Interaction

```{r height-quercus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_height_quercus$sem$plot)
```

#### SEM With Interaction

```{r height-quercus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_height_quercus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "growth",
  resp_var = "height",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_height_quercus_int$sem$plot)
```

---

## Growth: Diameter

### Timeseries (Beech and Oak)

```{r diameter-ts, fig.width=9.5, fig.height=5.8}
plot_ts_diam_inc(filter_soiltype = soil_type,
                 style = "band",
                 drought_bars = TRUE,
                 save_fig = FALSE, device = "pdf")
```

### Analysis Beech (Fagus)

#### Effect Size Over Time

```{r diameter-fagus-effect-size, fig.width=9.5, fig.height=5.8}
res_diameter_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "growth",
  resp_var = "diameter",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_diameter_fagus$temporal$plot)
```

#### SEM Without Interaction

```{r diameter-fagus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_diameter_fagus$sem$plot)
```

#### SEM With Interaction

```{r diameter-fagus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_diameter_fagus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "growth",
  resp_var = "diameter",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_diameter_fagus_int$sem$plot)
```

### Analysis Oak (Quercus)

#### Effect Size Over Time

```{r diameter-quercus-effect-size, fig.width=9.5, fig.height=5.8}
res_diameter_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "growth",
  resp_var = "diameter",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_diameter_quercus$temporal$plot)
```

#### SEM Without Interaction

```{r diameter-quercus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_diameter_quercus$sem$plot)
```

#### SEM With Interaction

```{r diameter-quercus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_diameter_quercus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "growth",
  resp_var = "diameter",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_diameter_quercus_int$sem$plot)
```

---

## Phenology: Leaf Development Stage

### Timeseries (Beech and Oak)

```{r phenology-ts, fig.width=9.5, fig.height=5.8}
plot_ts_phenology(style = "band",
                  filter_soiltype = soil_type,
                  save_fig = FALSE, device = "pdf")
```

### Analysis Beech (Fagus)

#### Effect Size Over Time

```{r phenology-fagus-effect-size, fig.width=9.5, fig.height=5.8}
res_phenology_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "phenology",
  resp_var = "stage",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_phenology_fagus$temporal$plot)
```

#### SEM Without Interaction

```{r phenology-fagus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_phenology_fagus$sem$plot)
```

#### SEM With Interaction

```{r phenology-fagus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_phenology_fagus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "phenology",
  resp_var = "stage",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_phenology_fagus_int$sem$plot)
```

### Analysis Oak (Quercus)

#### Effect Size Over Time

```{r phenology-quercus-effect-size, fig.width=9.5, fig.height=5.8}
res_phenology_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "phenology",
  resp_var = "stage",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_phenology_quercus$temporal$plot)
```

#### SEM Without Interaction

```{r phenology-quercus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_phenology_quercus$sem$plot)
```

#### SEM With Interaction

```{r phenology-quercus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_phenology_quercus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "phenology",
  resp_var = "stage",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_phenology_quercus_int$sem$plot)
```

---

## Quantum Yield (QY)

### Timeseries (Beech and Oak)

```{r qy-ts, fig.width=9.5, fig.height=5.8}
plot_ts_qy(filter_soiltype = soil_type,
           style = "band",
           drought_bars = TRUE,
           save_fig = FALSE, device = "pdf")
```

### Analysis Beech (Fagus)

#### Effect Size Over Time

```{r qy-fagus-effect-size, fig.width=9.5, fig.height=5.8}
res_qy_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "quantum_yield",
  resp_var = "qy",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_qy_fagus$temporal$plot)
```

#### SEM Without Interaction

```{r qy-fagus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_qy_fagus$sem$plot)
```

#### SEM With Interaction

```{r qy-fagus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_qy_fagus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "quantum_yield",
  resp_var = "qy",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_qy_fagus_int$sem$plot)
```

### Analysis Oak (Quercus)

#### Effect Size Over Time

```{r qy-quercus-effect-size, fig.width=9.5, fig.height=5.8}
res_qy_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "quantum_yield",
  resp_var = "qy",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_qy_quercus$temporal$plot)
```

#### SEM Without Interaction

```{r qy-quercus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_qy_quercus$sem$plot)
```

#### SEM With Interaction

```{r qy-quercus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_qy_quercus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "quantum_yield",
  resp_var = "qy",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_qy_quercus_int$sem$plot)
```

---

## Senescence: Percent Senesced

### Timeseries (Beech and Oak)

```{r senescence-pct-ts, fig.width=9.5, fig.height=5.8}
plot_ts_senescence_percent(filter_soiltype = soil_type,
                           style = "band",
                           drought_bars = FALSE,
                           save_fig = FALSE, device = "pdf")
```

### Analysis Beech (Fagus)

#### Effect Size Over Time

```{r senescence-pct-fagus-effect-size, fig.width=9.5, fig.height=5.8}
res_sen_pct_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "percent_senesced",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_sen_pct_fagus$temporal$plot)
```

#### SEM Without Interaction

```{r senescence-pct-fagus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_sen_pct_fagus$sem$plot)
```

#### SEM With Interaction

```{r senescence-pct-fagus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_sen_pct_fagus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "percent_senesced",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_sen_pct_fagus_int$sem$plot)
```

### Analysis Oak (Quercus)

#### Effect Size Over Time

```{r senescence-pct-quercus-effect-size, fig.width=9.5, fig.height=5.8}
res_sen_pct_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "percent_senesced",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_sen_pct_quercus$temporal$plot)
```

#### SEM Without Interaction

```{r senescence-pct-quercus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_sen_pct_quercus$sem$plot)
```

#### SEM With Interaction

```{r senescence-pct-quercus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_sen_pct_quercus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "percent_senesced",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_sen_pct_quercus_int$sem$plot)
```

---

## Senescence: Chlorophyll

### Timeseries (Beech and Oak)

```{r senescence-chl-ts, fig.width=9.5, fig.height=5.8}
plot_ts_senescence_chl(filter_soiltype = soil_type,
                       style = "band",
                       drought_bars = FALSE,
                       save_fig = FALSE, device = "pdf")
```

### Analysis Beech (Fagus)

#### Effect Size Over Time

```{r senescence-chl-fagus-effect-size, fig.width=9.5, fig.height=5.8}
res_sen_chl_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "chlavg",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_sen_chl_fagus$temporal$plot)
```

#### SEM Without Interaction

```{r senescence-chl-fagus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_sen_chl_fagus$sem$plot)
```

#### SEM With Interaction

```{r senescence-chl-fagus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_sen_chl_fagus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "chlavg",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_sen_chl_fagus_int$sem$plot)
```

### Analysis Oak (Quercus)

#### Effect Size Over Time

```{r senescence-chl-quercus-effect-size, fig.width=9.5, fig.height=5.8}
res_sen_chl_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "chlavg",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_sen_chl_quercus$temporal$plot)
```

#### SEM Without Interaction

```{r senescence-chl-quercus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_sen_chl_quercus$sem$plot)
```

#### SEM With Interaction

```{r senescence-chl-quercus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_sen_chl_quercus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "chlavg",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_sen_chl_quercus_int$sem$plot)
```

---

# SEM Aggregation Notebook

The revised SEM matrix workflow (treatments x target variables, separate direct/indirect/combined panels, plus SWC-related panels) is now maintained in a dedicated notebook:

- `notebooks/3-sem-aggregation.Rmd`

## Exploratory Phase-Specific SEM (Growth-Derived Targets)

Phase windows: until June, July-August, September-end.

```{r sem-safe-helper, include=FALSE}
safe_run_sem <- function(...) {
  tryCatch(run_sem_for_trait(...), error = function(e) {
    message("Skipping SEM fit: ", conditionMessage(e))
    NULL
  })
}
```

```{r sem-phase-growth-derived, fig.width=14, fig.height=5}
growth_targets <- c(
  "height_inc_t0_rel",
  "diameter_inc_t0_rel",
  "height_inc_phase_rel",
  "diameter_inc_phase_rel"
)

phase_windows <- c("until June", "July-August", "September-end")

for (sp in species_vec) {
  for (ph in phase_windows) {
    cat("\n\n### Species:", sp, "| phase:", ph, "\n")
    for (rv in growth_targets) {
      sem_phase <- safe_run_sem(
        type = "tree",
        data_name = "growth",
        resp_var = rv,
        species = sp,
        soil_type = soil_type,
        include_interaction = FALSE,
        phase_window = ph,
        scale_all_numeric = TRUE,
        swc_source = "measured",
        force_run = FALSE
      )

      if (is.null(sem_phase)) next

      cat("\n", rv, "| n:", sem_phase$n_rows_phase, "\n")
      print(sem_phase$matrix_plots$direct)
      print(sem_phase$matrix_plots$indirect)
      print(sem_phase$matrix_plots$total)
    }
  }
}
```

## Additional Growth-Derived Targets: Temporal + SEM

```{r growth-derived-temporal-sem, fig.width=9.5, fig.height=5.8}
safe_run_combo <- function(...) {
  tryCatch(make_temporal_sem_combo(...), error = function(e) {
    message("Skipping temporal+SEM combo: ", conditionMessage(e))
    NULL
  })
}

for (sp in species_vec) {
  for (rv in growth_targets) {
    cat("\n\n###", sp, "-", rv, "(no interaction)\n")
    res_no <- safe_run_combo(
      type = "tree",
      data_name = "growth",
      resp_var = rv,
      species = sp,
      soil_type = soil_type,
      include_interaction = FALSE,
      sem_phase_window = "all"
    )
    if (!is.null(res_no)) {
      print(res_no$temporal$plot)
      print(res_no$sem$plot)
    }

    cat("\n", sp, "-", rv, "(with interaction)\n")
    res_yes <- safe_run_combo(
      type = "tree",
      data_name = "growth",
      resp_var = rv,
      species = sp,
      soil_type = soil_type,
      include_interaction = TRUE,
      sem_phase_window = "all"
    )
    if (!is.null(res_yes)) {
      print(res_yes$sem$plot)
    }
  }
}
```

## Relative Growth Increments: Raw Timeseries

The following plots show treatment-level raw timeseries (mean ± SE) for the relative increment targets, analogous to the raw height and diameter timeseries charts.

```{r growth-relative-increment-timeseries, fig.width=10.5, fig.height=6.2}
plot_relative_growth_ts <- function(resp_var, y_label, panel_title) {
  df_rel <- get_data("tree", "growth") %>%
    dplyr::filter(!is.na(.data[[resp_var]]))

  if (!identical(soil_type, "both")) {
    df_rel <- df_rel %>% dplyr::filter(.data$soiltype == soil_type)
  }

  if (!nrow(df_rel)) {
    message("No rows available for ", resp_var, " after filtering.")
    return(invisible(NULL))
  }

  df_rel <- df_rel %>%
    dplyr::mutate(
      precipitation = factor(.data$precipitation, levels = c("control", "drought")),
      culture = factor(.data$culture, levels = c("mono", "mixed")),
      robinia = factor(.data$robinia, levels = c("without-robinia", "with-robinia")),
      species = factor(.data$species, levels = species_vec)
    )

  df_sum <- df_rel %>%
    dplyr::group_by(.data$species, .data$robinia, .data$precipitation, .data$culture, .data$date) %>%
    dplyr::summarise(
      mean_val = mean(.data[[resp_var]], na.rm = TRUE),
      se_val = sd(.data[[resp_var]], na.rm = TRUE) / sqrt(dplyr::n()),
      .groups = "drop"
    )

  p <- ggplot(
    df_sum,
    aes(
      x = date,
      y = mean_val,
      color = precipitation,
      fill = precipitation,
      linetype = culture,
      group = interaction(precipitation, culture)
    )
  ) +
    geom_ribbon(aes(ymin = mean_val - se_val, ymax = mean_val + se_val), alpha = 0.16, color = NA) +
    geom_line(linewidth = 0.9) +
    facet_grid(species ~ robinia) +
    scale_color_manual(values = c(control = "grey50", drought = "indianred")) +
    scale_fill_manual(values = c(control = "grey50", drought = "indianred")) +
    scale_linetype_manual(values = c(mono = "solid", mixed = "dotted")) +
    labs(
      title = panel_title,
      x = "Date",
      y = y_label,
      color = "Precipitation",
      fill = "Precipitation",
      linetype = "Culture"
    ) +
    theme_bw(base_size = 11) +
    theme(
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 35, hjust = 1),
      legend.position = "bottom"
    )

  print(p)
}

plot_relative_growth_ts(
  resp_var = "height_inc_t0_rel",
  y_label = "Relative height increment from first measurement (-)",
  panel_title = "Height relative increment (baseline = first measurement)"
)

plot_relative_growth_ts(
  resp_var = "diameter_inc_t0_rel",
  y_label = "Relative diameter increment from first measurement (-)",
  panel_title = "Diameter relative increment (baseline = first measurement)"
)

plot_relative_growth_ts(
  resp_var = "height_inc_phase_rel",
  y_label = "Relative height increment within phase (-)",
  panel_title = "Height relative increment (chained phase baseline)"
)

plot_relative_growth_ts(
  resp_var = "diameter_inc_phase_rel",
  y_label = "Relative diameter increment within phase (-)",
  panel_title = "Diameter relative increment (chained phase baseline)"
)
```

---

# Session Info

```{r session-info}
sessionInfo()
```
