---
title: 'ALINV – SWC Interpolation: GAM Imputation & SEM Sensitivity'
date: "`r format(Sys.Date(), '%d %b %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    highlight: espresso
    theme: journal
    fig_width: 9
    fig_height: 6
execute:
  warning: false
  message: false
  echo: true
  results: hide
---

```{r setup, include=FALSE}
# Knitr settings
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  results = 'hide',
  fig.show = 'hold',
  fig.width = 9,
  fig.height = 6
)

# Resolve project root from notebook location (robust to knit working directory)
input_file <- knitr::current_input()
input_dir <- dirname(normalizePath(input_file, winslash = "/", mustWork = TRUE))
proj_root <- normalizePath(file.path(input_dir, ".."), winslash = "/", mustWork = TRUE)

# Function libraries
source(file.path(proj_root, "functions", "_source.R"))
source(file.path(proj_root, "functions", "1-summary-figures.R"))
source(file.path(proj_root, "functions", "3-effect-size-factorial.R"))
source(file.path(proj_root, "functions", "4-structural-equation-model.R"))
source(file.path(proj_root, "functions", "5-combining-glmm-sem.R"))

library(tidyverse)
library(mgcv)
library(lubridate)
library(patchwork)

# Global choices
soil_type <- "both"
species_vec <- c("fagus", "quercus")
```

# Introduction

Soil water content (SWC) is measured manually at irregular intervals in the ALINV experiment. The piecewise SEMs in the main analysis notebook use observed SWC matched to tree measurement dates within a ±7-day window. This notebook provides:

1. **GAM Imputation** — a treatment-agnostic generalised additive model that produces daily SWC estimates from meteorological and soil-water-potential predictors, ensuring imputed values carry no treatment information.
2. **SEM Sensitivity Comparison** — systematic re-fitting of all trait-level SEMs using the GAM-imputed SWC in place of the measured values, to assess whether the choice of SWC source materially changes structural conclusions.

## Methods

### SWC Imputation GAM

A treatment-agnostic GAM imputes daily SWC from site-level meteorological drivers:

$$\text{SWC}_{it} = f_1(t) + f_2(\text{SWP}_{\text{site},t}) + f_3(T_{\text{air},t}) + f_4(\text{VPD}_t) + f_5(\text{Rad}_t) + b_i + \varepsilon_{it}$$

where $f_1, \ldots, f_5$ are penalised thin-plate splines and $b_i$ is a random intercept per box. Because only site-level predictors enter the model, the imputed SWC does not encode treatment group information.

### SEM Sensitivity

Each piecewise SEM from the main analysis is re-fitted with GAM-imputed SWC replacing the observed values. R² metrics and path coefficients are compared side by side to identify any substantive changes in direct vs. indirect treatment effect decomposition.

---

# SWC Imputation: Treatment-Agnostic GAM

## Data Preparation

```{r gam-meteo-daily}
meteo_file <- .resolve_path("data/raw/sensor_data/meteo_10min.dat")
meteo_raw <- read.table(
  meteo_file, sep = ",", header = FALSE, skip = 4,
  stringsAsFactors = FALSE,
  col.names = c("datetime", "id", "air_temp", "rh", "patm", "vpd", "radiation")
)

meteo_daily <- meteo_raw %>%
  mutate(datetime = ymd_hms(datetime), date = as.Date(datetime)) %>%
  mutate(across(c(air_temp, rh, vpd, radiation), as.numeric)) %>%
  group_by(date) %>%
  summarise(
    air_temp = mean(air_temp, na.rm = TRUE),
    rh       = mean(rh, na.rm = TRUE),
    vpd      = mean(vpd, na.rm = TRUE),
    radiation = mean(radiation, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r gam-swp-daily}
soil_file <- .resolve_path("data/raw/sensor_data/soil_10min.dat")
soil_raw <- read.table(
  soil_file, sep = ",", header = TRUE, skip = 2, stringsAsFactors = FALSE
) %>% slice(-1)

swp_site_daily <- soil_raw %>%
  mutate(datetime = ymd_hms(TS), date = as.Date(datetime)) %>%
  select(date, starts_with("kPa")) %>%
  mutate(across(-date, ~ as.numeric(na_if(.x, "NAN")))) %>%
  group_by(date) %>%
  summarise(across(everything(), \(x) mean(x, na.rm = TRUE)), .groups = "drop") %>%
  pivot_longer(starts_with("kPa"), names_to = "sensor_raw", values_to = "soil_mp") %>%
  filter(!is.na(soil_mp), soil_mp < 0) %>%
  mutate(swp_log10 = log10(-soil_mp)) %>%
  group_by(date) %>%
  summarise(swp_site = median(swp_log10, na.rm = TRUE), .groups = "drop")
```

```{r gam-panel-build}
swc_obs <- readr::read_csv(
  .resolve_path("data/interim/box_soilwater.csv"), show_col_types = FALSE
) %>%
  transmute(boxlabel = as.character(boxlabel), date = as.Date(date), swc_obs = as.numeric(swc))

date_min <- max(min(meteo_daily$date), min(swp_site_daily$date))
date_max <- min(max(meteo_daily$date), max(swp_site_daily$date))

panel_daily <- tidyr::expand_grid(
  swc_obs %>% distinct(boxlabel),
  tibble(date = seq.Date(date_min, date_max, by = "day"))
) %>%
  left_join(swc_obs, by = c("boxlabel", "date")) %>%
  left_join(meteo_daily, by = "date") %>%
  left_join(swp_site_daily, by = "date") %>%
  mutate(date_num = as.numeric(date), boxlabel = factor(boxlabel))

fit_data <- panel_daily %>%
  filter(!is.na(swc_obs), !is.na(swp_site), !is.na(air_temp), !is.na(vpd), !is.na(radiation))
```

## GAM Fit & Diagnostics

```{r gam-fit}
gam_swc <- mgcv::gam(
  swc_obs ~ s(date_num, k = 10) + s(swp_site, k = 8) +
    s(air_temp, k = 8) + s(vpd, k = 8) + s(radiation, k = 8) +
    s(boxlabel, bs = "re"),
  data = fit_data, method = "REML"
)
```

```{r gam-summary, results='markup'}
summary(gam_swc)
```

```{r gam-smooth-plots, fig.width=11, fig.height=8}
plot(gam_swc, pages = 1, shade = TRUE, residuals = TRUE, scale = 0)
```

## Prediction & Validation

```{r gam-predict}
pred_data <- panel_daily %>%
  filter(!is.na(swp_site), !is.na(air_temp), !is.na(vpd), !is.na(radiation))
pred <- predict(gam_swc, newdata = pred_data, se.fit = TRUE, type = "response")

gam_out <- pred_data %>%
  mutate(
    swc_hat    = pmin(pmax(as.numeric(pred$fit), 0), 100),
    swc_hat_se = as.numeric(pred$se.fit),
    swc_hat_lo = pmax(swc_hat - 1.96 * swc_hat_se, 0),
    swc_hat_hi = pmin(swc_hat + 1.96 * swc_hat_se, 100),
    swc_source = if_else(!is.na(swc_obs), "measured", "imputed_gam")
  ) %>%
  arrange(boxlabel, date)
```

```{r gam-metrics, results='markup'}
gam_out %>%
  filter(!is.na(swc_obs)) %>%
  summarise(n = n(),
            rmse = sqrt(mean((swc_obs - swc_hat)^2)),
            mae  = mean(abs(swc_obs - swc_hat)),
            r    = cor(swc_obs, swc_hat))
```

```{r gam-obs-vs-fit}
gam_out %>%
  filter(!is.na(swc_obs)) %>%
  ggplot(aes(swc_obs, swc_hat)) +
  geom_point(alpha = 0.35, size = 1.1, color = "grey25") +
  geom_abline(slope = 1, intercept = 0, color = "firebrick", linewidth = 0.7) +
  coord_equal() +
  labs(title = "Observed vs Fitted SWC", x = "Observed SWC", y = "Fitted SWC") +
  theme_linedraw()
```

```{r gam-residuals}
gam_res <- gam_out %>% filter(!is.na(swc_obs)) %>% mutate(resid = swc_obs - swc_hat)

p_res1 <- ggplot(gam_res, aes(swc_hat, resid)) +
  geom_point(alpha = 0.35, size = 1.0) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "firebrick") +
  labs(title = "Residuals vs Fitted", x = "Fitted SWC", y = "Residual") +
  theme_linedraw()

p_res2 <- ggplot(gam_res, aes(resid)) +
  geom_histogram(bins = 35, fill = "grey70", color = "white") +
  labs(title = "Residual Distribution", x = "Residual", y = "Count") +
  theme_linedraw()

p_res1
p_res2
```

## Example Box Trajectories

```{r gam-trajectories, fig.width=11, fig.height=7}
set.seed(42)
example_boxes <- gam_out %>% distinct(boxlabel) %>% slice_sample(n = 6) %>% pull(boxlabel)

gam_out %>%
  filter(boxlabel %in% example_boxes) %>%
  ggplot(aes(date, swc_hat)) +
  geom_ribbon(aes(ymin = swc_hat_lo, ymax = swc_hat_hi), fill = "grey80", alpha = 0.5) +
  geom_line(color = "steelblue", linewidth = 0.5) +
  geom_point(data = ~ dplyr::filter(.x, !is.na(swc_obs)), aes(y = swc_obs),
             color = "black", size = 1.1, alpha = 0.8) +
  facet_wrap(~ boxlabel, scales = "free_y", ncol = 3) +
  labs(title = "Daily SWC Trajectories (Lines = GAM, Points = Measured)", x = "Date", y = "SWC") +
  theme_linedraw()
```

## Export Imputed SWC

```{r gam-export}
readr::write_csv(gam_out, .resolve_path("data/interim/box_soilwater_daily_gam_agnostic.csv"))
```

---

# SWC Source Comparison: Influence on SEM Results

The SEMs in the main analysis notebook used measured SWC. Here we systematically compare SEM results when using GAM-imputed SWC instead, to assess whether the choice of SWC source materially changes conclusions.

```{r swc-comp-controls}
swc_levels <- c("measured", "imputed_gam")
species_levels <- c("fagus", "quercus")
interaction_levels <- c(FALSE, TRUE)

swc_model_grid <- tribble(
  ~section, ~data_name, ~resp_var,
  "Chlorophyll",            "chlorophyll",  "chl",
  "Tree Condition",         "condition",    "condition",
  "Growth Height",          "growth",       "height",
  "Growth Diameter",        "growth",       "diameter",
  "Phenology Stage",        "phenology",    "stage",
  "Quantum Yield",          "quantum_yield","qy",
  "Senescence Percent",     "senescence",   "percent_senesced",
  "Senescence Chlorophyll", "senescence",   "chlavg"
)
```

```{r swc-comp-helpers}
fit_sem_swc <- function(data_name, resp_var, species, include_interaction, swc_source) {
  tryCatch(
    run_sem_for_trait(
      type = "tree", data_name = data_name, resp_var = resp_var,
      species = species, soil_type = soil_type,
      include_interaction = include_interaction,
      scale_all_numeric = TRUE, do_rfe = FALSE,
      swc_source = swc_source, force_run = FALSE
    ),
    error = function(e) { message("SEM fit failed: ", conditionMessage(e)); NULL }
  )
}

plot_sem_pair <- function(res_meas, res_imp, panel_title = NULL) {
  p_meas <- if (!is.null(res_meas) && !is.null(res_meas$plot)) {
    res_meas$plot + ggtitle("Measured SWC")
  } else {
    ggplot() + theme_void() + ggtitle("Measured SWC (missing)")
  }
  p_imp <- if (!is.null(res_imp) && !is.null(res_imp$plot)) {
    res_imp$plot + ggtitle("Imputed SWC")
  } else {
    ggplot() + theme_void() + ggtitle("Imputed SWC (missing)")
  }
  p <- p_meas | p_imp
  if (!is.null(panel_title)) p <- p + plot_annotation(title = panel_title)
  p
}

extract_sem_r2 <- function(res) {
  if (is.null(res) || is.null(res$sem)) {
    return(tibble(sem_r2_swc_marginal = NA_real_, sem_r2_swc_conditional = NA_real_,
                  sem_r2_resp_marginal = NA_real_, sem_r2_resp_conditional = NA_real_))
  }
  r2_tbl <- tryCatch(piecewiseSEM::rsquared(res$sem), error = function(e) NULL)
  if (is.null(r2_tbl)) {
    return(tibble(sem_r2_swc_marginal = NA_real_, sem_r2_swc_conditional = NA_real_,
                  sem_r2_resp_marginal = NA_real_, sem_r2_resp_conditional = NA_real_))
  }
  tibble(
    sem_r2_swc_marginal    = r2_tbl$Marginal[r2_tbl$Response == "swc"][1],
    sem_r2_swc_conditional = r2_tbl$Conditional[r2_tbl$Response == "swc"][1],
    sem_r2_resp_marginal   = r2_tbl$Marginal[r2_tbl$Response == "y"][1],
    sem_r2_resp_conditional = r2_tbl$Conditional[r2_tbl$Response == "y"][1]
  )
}

extract_sem_paths <- function(res) {
  empty <- tibble(factor = character(), effect_class = character(), path = character(),
                  estimate = double(), p_value = double(), abs_estimate = double())
  if (is.null(res)) return(empty)
  out <- list()
  if (!is.null(res$effects) && nrow(res$effects) > 0) {
    out[[length(out) + 1L]] <- bind_rows(
      res$effects %>% transmute(factor, effect_class = "main", path = "to_swc (a)", estimate = a, p_value = p_a),
      res$effects %>% transmute(factor, effect_class = "main", path = "to_target_direct (c')", estimate = c_direct, p_value = p_c),
      res$effects %>% transmute(factor, effect_class = "main", path = "to_target_indirect (a*b)", estimate = indirect, p_value = p_ind),
      res$effects %>% transmute(factor, effect_class = "main", path = "to_target_total", estimate = total, p_value = p_tot)
    ) %>% mutate(abs_estimate = abs(estimate))
  }
  if (!is.null(res$effects_int) && nrow(res$effects_int) > 0) {
    out[[length(out) + 1L]] <- bind_rows(
      res$effects_int %>% transmute(factor, effect_class = "interaction", path = "to_swc (a)", estimate = a, p_value = p_a),
      res$effects_int %>% transmute(factor, effect_class = "interaction", path = "to_target_direct (c')", estimate = c_direct, p_value = p_c),
      res$effects_int %>% transmute(factor, effect_class = "interaction", path = "to_target_indirect (a*b)", estimate = indirect, p_value = p_ind),
      res$effects_int %>% transmute(factor, effect_class = "interaction", path = "to_target_total", estimate = total, p_value = p_tot)
    ) %>% mutate(abs_estimate = abs(estimate))
  }
  if (!length(out)) return(empty)
  bind_rows(out)
}
```

## Fit SEM Grid (Measured vs Imputed)

```{r swc-comp-fit-all}
swc_fit_tasks <- tidyr::crossing(
  swc_model_grid,
  species = species_levels,
  include_interaction = interaction_levels,
  swc_source = swc_levels
) %>%
  mutate(result = purrr::pmap(
    list(data_name, resp_var, species, include_interaction, swc_source), fit_sem_swc
  ))

swc_metrics_long <- swc_fit_tasks %>%
  mutate(metrics = purrr::map(result, extract_sem_r2)) %>%
  tidyr::unnest(metrics) %>%
  mutate(target_label = paste0(section, " | ", species, if_else(include_interaction, " | int", " | no-int")))

swc_sem_paths_long <- swc_fit_tasks %>%
  mutate(sem_paths = purrr::map(result, extract_sem_paths)) %>%
  tidyr::unnest(sem_paths)
```

## Faceted 1:1 SEM Metrics

```{r swc-comp-1to1-metrics, fig.width=14, fig.height=10}
sem_metric_cols <- c("sem_r2_swc_marginal", "sem_r2_swc_conditional",
                     "sem_r2_resp_marginal", "sem_r2_resp_conditional")

swc_metrics_wide <- swc_metrics_long %>%
  select(section, data_name, resp_var, species, include_interaction, target_label,
         swc_source, all_of(sem_metric_cols)) %>%
  pivot_longer(cols = all_of(sem_metric_cols), names_to = "metric", values_to = "value") %>%
  pivot_wider(names_from = swc_source, values_from = value) %>%
  mutate(delta_imputed_minus_measured = imputed_gam - measured)

metrics_plot_df <- swc_metrics_wide %>%
  filter(is.finite(measured), is.finite(imputed_gam)) %>%
  mutate(
    metric = factor(metric,
      levels = sem_metric_cols,
      labels = c("SWC R2 Marginal", "SWC R2 Conditional",
                 "Target R2 Marginal", "Target R2 Conditional")),
    interaction_label = if_else(include_interaction, "With interaction", "No interaction")
  )

ggplot(metrics_plot_df, aes(x = measured, y = imputed_gam, color = species)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey40") +
  geom_point(size = 2.2, alpha = 0.9) +
  geom_text(aes(label = section), size = 2.8, check_overlap = TRUE, vjust = -0.6, show.legend = FALSE) +
  facet_grid(interaction_label ~ metric) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(title = "SEM performance by SWC source",
       subtitle = "Points above 1:1 line indicate higher metric with imputed SWC",
       x = "Measured SWC", y = "Imputed SWC", color = "Species") +
  theme_linedraw(base_size = 11) + theme(legend.position = "bottom")
```

## Factor Influence: Measured vs Imputed SWC

```{r swc-comp-factor-influence, fig.width=16, fig.height=10}
infl_df <- swc_sem_paths_long %>%
  filter(include_interaction == FALSE, effect_class == "main") %>%
  filter(path %in% c("to_swc (a)", "to_target_direct (c')", "to_target_indirect (a*b)", "to_target_total"))

summarise_importance <- function(df, species_keep = NULL) {
  x <- df
  if (!is.null(species_keep)) x <- x %>% filter(species == species_keep)
  x %>%
    group_by(swc_source, factor, path) %>%
    summarise(mean_abs = mean(abs_estimate, na.rm = TRUE),
              se_abs = sd(abs_estimate, na.rm = TRUE) / sqrt(dplyr::n()),
              .groups = "drop")
}

plot_importance_panel <- function(df_imp, title_txt) {
  ggplot(df_imp, aes(x = reorder(factor, mean_abs), y = mean_abs, fill = swc_source)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.72) +
    geom_errorbar(aes(ymin = pmax(0, mean_abs - se_abs), ymax = mean_abs + se_abs),
                  position = position_dodge(width = 0.8), width = 0.2) +
    facet_wrap(~ path, scales = "free_y", ncol = 2) + coord_flip() +
    labs(title = title_txt, x = "Factor", y = "Mean |effect|", fill = "SWC source") +
    theme_linedraw(base_size = 11) + theme(legend.position = "bottom")
}

p_all <- plot_importance_panel(summarise_importance(infl_df), "All species")
p_fag <- plot_importance_panel(summarise_importance(infl_df, "fagus"), "Fagus")
p_que <- plot_importance_panel(summarise_importance(infl_df, "quercus"), "Quercus")

(p_all | p_fag | p_que) +
  patchwork::plot_layout(guides = "collect") +
  patchwork::plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom", plot.tag = element_text(face = "bold"))
```

## Full SEM Figure Comparison

```{r swc-comp-sem-figures, results='asis', fig.width=16, fig.height=7}
get_swc_fit <- function(sec, dn, rv, sp, int, src) {
  swc_fit_tasks %>%
    filter(.data$section == sec, .data$data_name == dn, .data$resp_var == rv,
           .data$species == sp, .data$include_interaction == int, .data$swc_source == src) %>%
    pull(result) %>% .[[1]]
}

for (i in seq_len(nrow(swc_model_grid))) {
  row <- swc_model_grid[i, ]
  cat("\n\n### ", row$section, "\n", sep = "")
  for (sp in species_levels) {
    cat("\n\n#### Species: ", sp, "\n", sep = "")

    res_meas <- get_swc_fit(row$section, row$data_name, row$resp_var, sp, FALSE, "measured")
    res_imp  <- get_swc_fit(row$section, row$data_name, row$resp_var, sp, FALSE, "imputed_gam")
    cat("\n\n##### No Interaction\n")
    print(plot_sem_pair(res_meas, res_imp,
      paste0("SEM no interaction: ", row$data_name, "/", row$resp_var, "/", sp)))

    res_meas_int <- get_swc_fit(row$section, row$data_name, row$resp_var, sp, TRUE, "measured")
    res_imp_int  <- get_swc_fit(row$section, row$data_name, row$resp_var, sp, TRUE, "imputed_gam")
    cat("\n\n##### With Interaction\n")
    print(plot_sem_pair(res_meas_int, res_imp_int,
      paste0("SEM with interaction: ", row$data_name, "/", row$resp_var, "/", sp)))
  }
}
```

---

# Session Info

```{r session-info}
sessionInfo()
```
