---
title: 'ALINV – Comprehensive Analysis: Timeseries & Modeling Results'
date: "`r format(Sys.Date(), '%d %b %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    highlight: espresso
    theme: journal
    fig_width: 9
    fig_height: 6
execute:
  warning: false
  message: false
  echo: true
  results: hide
---

```{r setup, include=FALSE}
# Knitr settings
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  results = 'hide',
  fig.show = 'hold',
  fig.width = 9,
  fig.height = 6
)

# Resolve project root from notebook location (robust to knit working directory)
input_file <- knitr::current_input()
input_dir <- dirname(normalizePath(input_file, winslash = "/", mustWork = TRUE))
proj_root <- normalizePath(file.path(input_dir, ".."), winslash = "/", mustWork = TRUE)

# Function libraries
source(file.path(proj_root, "functions", "_source.R"))
source(file.path(proj_root, "functions", "0-summary_figures.R"))
source(file.path(proj_root, "functions", "0-biomass.R"))
source(file.path(proj_root, "functions", "x-effect-size-factorial.R"))
source(file.path(proj_root, "functions", "x-structural-equation-model.R"))
source(file.path(proj_root, "functions", "x-effect-size-model-aggregation.R"))
source(file.path(proj_root, "functions", "x-combining-glmm-sem.R"))

library(mgcv)
library(lubridate)
library(patchwork)

# Global choices
soil_type <- "both"        # "both", "inoc-beech", or "inoc-robinia"
species_vec <- c("fagus", "quercus")
```

# Introduction

This notebook presents a comprehensive analysis of the ALINV (Alien Invasion) experiment, which investigates how drought, species mixing, *Robinia* presence, and legacy soil type affect tree physiology and growth. The experiment uses a full-factorial design with two target species — European beech (*Fagus sylvatica*) and sessile oak (*Quercus petraea*) — grown in mesocosm boxes under controlled treatment combinations.

**Contents:**

1. **Treatment Effects Summary** — aggregated temporal GLMM effect sizes across all response variables, grouped by experimental phase.
2. **Biomass** — end-of-experiment destructive harvest data (shoot, root, root:shoot ratio), visualised as treatment boxplots and quantified with a one-off GLMM.
3. **Soil Conditions** — time series of soil water content (SWC), soil water potential (SWP), soil respiration, and C/N isotopes.
4. **Tree Variables** — for each trait (chlorophyll, condition, height growth, diameter growth, phenology, quantum yield, senescence), the notebook shows: (i) a treatment-level time series, (ii) temporal effect sizes from date-specific GLMMs, and (iii) piecewise structural equation models (SEM) decomposing direct and SWC-mediated indirect effects.
5. **SWC Imputation** — a treatment-agnostic GAM imputing daily SWC from climate and soil water potential data, with diagnostics.
6. **SWC Source Comparison** — systematic comparison of SEM results using measured vs. GAM-imputed SWC, assessing sensitivity of conclusions to the SWC source.

## Methods

### Temporal GLMM (time-varying treatment effects)

For each tree-level response variable, a linear mixed-effects model is fitted across all measurement dates:

$$y_{ijk} = \mu + \alpha_t + (\alpha_t \times \text{Robinia})_{j} + (\alpha_t \times \text{Precipitation})_{j} + (\alpha_t \times \text{Culture})_{j} + (\alpha_t \times \text{Soiltype})_{j} + b_{\text{box}} + b_{\text{tree}} + \varepsilon_{ijk}$$

where $\alpha_t$ is a date fixed effect and treatment effects are estimated per date via date×treatment interactions. Random intercepts are fitted for box ($b_{\text{box}}$) and tree ($b_{\text{tree}}$). Treatment contrasts at each date are extracted using estimated marginal means (`emmeans`, reverse pairwise).

### Piecewise Structural Equation Model (SEM)

Two-equation piecewise SEM fitted separately per species and trait, using `piecewiseSEM::psem()`:

**SWC sub-model:**
$$\text{SWC} = \beta_0 + \beta_1 \text{DOY} + \beta_2 \text{DOY}^2 + \beta_3 \text{Precipitation} + \beta_4 \text{Robinia} + \beta_5 \text{Soiltype} + \beta_6 \text{Culture} + \beta_7 \text{ExtremeEvent} + b_{\text{box}} + \varepsilon$$

**Response sub-model:**
$$y = \gamma_0 + \gamma_1 \text{DOY} + \gamma_2 \text{DOY}^2 + \gamma_3 \text{SWC} + \gamma_4 \text{Precipitation} + \cdots + b_{\text{box}} + b_{\text{tree}} + \varepsilon$$

Both are `lme4::lmer()` models. Treatment effects are decomposed into:

- **Direct path** ($c'$): treatment → response
- **Indirect path** ($a \times b$): treatment → SWC → response
- **Total effect**: $c' + a \times b$

All numeric predictors are centred and scaled before fitting.

### Biomass GLMM

A single-timepoint linear mixed model for each biomass metric (shoot, root, root:shoot ratio), fitted on z-scored responses to produce standardised coefficients:

$$y_z = \beta_0 + \beta_1 \text{Precipitation} + \beta_2 \text{Culture} + \beta_3 \text{Robinia} + \beta_4 \text{Soiltype} + b_{\text{box}} + \varepsilon$$

Coefficients are in standard deviation units of the response. 95% Wald confidence intervals are used to assess significance.

### SWC Imputation GAM

A treatment-agnostic GAM fitted with `mgcv::gam()` (REML) to produce daily SWC predictions:

$$\text{SWC}_{it} = f_1(t) + f_2(\text{SWP}_{\text{site},t}) + f_3(T_{\text{air},t}) + f_4(\text{VPD}_t) + f_5(\text{Rad}_t) + b_i + \varepsilon_{it}$$

where $f_1, \ldots, f_5$ are thin-plate regression splines and $b_i \sim N(0, \sigma^2_b)$ is a random intercept per box. No treatment variables enter the model, ensuring imputed SWC is treatment-agnostic.

---

# Overview: Treatment Effects Summary

Aggregated temporal GLMM effect sizes across all tree response variables, grouped by experimental phase (until June / July–August / September+).

```{r treatment-effects-overview, fig.width=6, fig.height=12}
# Collect all temporal GLMM effects across variables and species
all_eff <- collect_all_effects(
  var_grid    = get_data_var_grid() %>% dplyr::filter(type == "tree"),
  species_vec = c("fagus", "quercus"),
  soil_type   = "both"
)

if (nrow(all_eff) > 0) {
  all_eff_imp <- all_eff %>% 
    add_importance_metric(use_stat = FALSE, scale_within = TRUE)
  all_eff_imp_phase <- add_phase_to_effects(all_eff_imp)
  
  # Overall effects across all species
  p_all <- plot_treatment_importance_phases(
    all_eff_imp_phase, 
    stat = "mean_se"
  ) + coord_cartesian(ylim = c(0, 3)) + ggtitle("All species")
  
  # Fagus only
  p_fag <- plot_treatment_importance_phases(
    all_eff_imp_phase %>% dplyr::filter(species == "fagus"),
    stat = "mean_se"
  ) + coord_cartesian(ylim = c(0, 3)) + ggtitle("Fagus")
  
  # Quercus only
  p_que <- plot_treatment_importance_phases(
    all_eff_imp_phase %>% dplyr::filter(species == "quercus"),
    stat = "mean_se"
  ) + coord_cartesian(ylim = c(0, 3)) + ggtitle("Quercus")
  
  p_combined <- (p_all / p_fag / p_que) +
    patchwork::plot_layout(guides = "collect") +
    patchwork::plot_annotation(tag_levels = "A") &
    ggplot2::theme(legend.position = "bottom",
                   plot.tag = ggplot2::element_text(face = "bold"))
  
  print(p_combined)
} else {
  message("No effects data collected. This may happen on first run.")
}
```

---

# Biomass: Treatment Comparison

```{r biomass-data}
df_biomass <- wrangle_tree_biomass(file.path(proj_root, "data", "raw", "data_20260309.xlsx"))
```

## Fagus – Boxplots

```{r biomass-boxplot-fagus, fig.width=9, fig.height=10}
print(plot_tree_biomass_treatments(df_biomass, species_keep = "fagus"))
```

## Quercus – Boxplots

```{r biomass-boxplot-quercus, fig.width=9, fig.height=10}
print(plot_tree_biomass_treatments(df_biomass, species_keep = "quercus"))
```

## GLMM: Treatment Effects on Biomass

A one-off GLMM (no temporal factor) quantifies the influence of precipitation, culture, robinia presence, and soil type on root, shoot, and root:shoot biomass.

### Fagus

```{r biomass-glmm-fagus, fig.width=8, fig.height=6}
res_fag <- run_biomass_glmm_species(df_biomass, species_keep = "fagus")
print(res_fag$plot)
```

```{r biomass-glmm-fagus-perf, results='asis'}
knitr::kable(res_fag$performance %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))),
  caption = "Fagus biomass GLMM performance")
```

### Quercus

```{r biomass-glmm-quercus, fig.width=8, fig.height=6}
res_que <- run_biomass_glmm_species(df_biomass, species_keep = "quercus")
print(res_que$plot)
```

```{r biomass-glmm-quercus-perf, results='asis'}
knitr::kable(res_que$performance %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))),
  caption = "Quercus biomass GLMM performance")
```

---

# Context: Soil Conditions

## Soil Water Content (SWC)

```{r swc, fig.width=12, fig.height=6}
plot_ts_swc(filter_soiltype = soil_type,
            style = "band",
            drought_bars = TRUE,
            save_fig = FALSE, device = "pdf")
```

## Soil Water Potential (Matric Potential)

### Raw Data + LOESS Smoothed
```{r water-potential-both, fig.width=10, fig.height=6}
plot_ts_soil_mp(
  filter_soiltype = soil_type,
  show="both",
  save_fig = FALSE,
  device = "pdf"
)
```

### Raw Data
```{r water-potential-raw, fig.width=10, fig.height=6}
plot_ts_soil_mp(
  filter_soiltype = soil_type,
  show="raw",
  save_fig = FALSE,
  device = "pdf"
)
```


## Soil Respiration

```{r respiration, fig.width=12, fig.height=6}
plot_ts_resp(filter_soiltype = soil_type,
            style = "band",
            drought_bars = TRUE,
            save_fig = FALSE, device = "pdf")
```


## Soil C/N & Isotopes

```{r isotopes, fig.width=12, fig.height=10}
boxplot_soil_cn_isotopes(save_fig = FALSE, device = "pdf")
```

---

# Tree Variables: Timeseries, Temporal Models & SEM

## Chlorophyll Content (Chl)

### Timeseries (Beech and Oak)

```{r chl-ts, fig.width=9.5, fig.height=5.8}
plot_ts_chl(filter_soiltype = soil_type,
            style = "band",
            drought_bars = TRUE,
            save_fig = FALSE, device = "pdf")
```

### Analysis Beech (Fagus)

#### Effect Size Over Time

```{r chl-fagus-effect-size, fig.width=9.5, fig.height=5.8}
res_chl_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "chlorophyll",
  resp_var = "chl",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_chl_fagus$temporal$plot)
```

#### SEM Without Interaction

```{r chl-fagus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_chl_fagus$sem$plot)
```

#### SEM With Interaction

```{r chl-fagus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_chl_fagus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "chlorophyll",
  resp_var = "chl",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_chl_fagus_int$sem$plot)
```

### Analysis Oak (Quercus)

#### Effect Size Over Time

```{r chl-quercus-effect-size, fig.width=9.5, fig.height=5.8}
res_chl_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "chlorophyll",
  resp_var = "chl",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_chl_quercus$temporal$plot)
```

#### SEM Without Interaction

```{r chl-quercus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_chl_quercus$sem$plot)
```

#### SEM With Interaction

```{r chl-quercus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_chl_quercus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "chlorophyll",
  resp_var = "chl",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_chl_quercus_int$sem$plot)
```

---

## Tree Condition

### Timeseries (Beech and Oak)

```{r cond-ts, fig.width=9.5, fig.height=5.8}
plot_ts_condition(filter_soiltype = soil_type,
                  style = "band",
                  drought_bars = TRUE,
                  save_fig = FALSE, device = "pdf")
```

### Analysis Beech (Fagus)

#### Effect Size Over Time

```{r cond-fagus-effect-size, fig.width=9.5, fig.height=5.8}
res_cond_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "condition",
  resp_var = "condition",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_cond_fagus$temporal$plot)
```

#### SEM Without Interaction

```{r cond-fagus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_cond_fagus$sem$plot)
```

#### SEM With Interaction

```{r cond-fagus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_cond_fagus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "condition",
  resp_var = "condition",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_cond_fagus_int$sem$plot)
```

### Analysis Oak (Quercus)

#### Effect Size Over Time

```{r cond-quercus-effect-size, fig.width=9.5, fig.height=5.8}
res_cond_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "condition",
  resp_var = "condition",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_cond_quercus$temporal$plot)
```

#### SEM Without Interaction

```{r cond-quercus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_cond_quercus$sem$plot)
```

#### SEM With Interaction

```{r cond-quercus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_cond_quercus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "condition",
  resp_var = "condition",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_cond_quercus_int$sem$plot)
```

---

## Growth: Height

### Timeseries (Beech and Oak)

```{r height-ts, fig.width=9.5, fig.height=5.8}
plot_ts_height_inc(filter_soiltype = soil_type,
                   style = "band",
                   drought_bars = TRUE,
                   save_fig = FALSE, device = "pdf")
```

### Analysis Beech (Fagus)

#### Effect Size Over Time

```{r height-fagus-effect-size, fig.width=9.5, fig.height=5.8}
res_height_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "growth",
  resp_var = "height",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_height_fagus$temporal$plot)
```

#### SEM Without Interaction

```{r height-fagus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_height_fagus$sem$plot)
```

#### SEM With Interaction

```{r height-fagus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_height_fagus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "growth",
  resp_var = "height",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_height_fagus_int$sem$plot)
```

### Analysis Oak (Quercus)

#### Effect Size Over Time

```{r height-quercus-effect-size, fig.width=9.5, fig.height=5.8}
res_height_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "growth",
  resp_var = "height",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_height_quercus$temporal$plot)
```

#### SEM Without Interaction

```{r height-quercus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_height_quercus$sem$plot)
```

#### SEM With Interaction

```{r height-quercus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_height_quercus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "growth",
  resp_var = "height",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_height_quercus_int$sem$plot)
```

---

## Growth: Diameter

### Timeseries (Beech and Oak)

```{r diameter-ts, fig.width=9.5, fig.height=5.8}
plot_ts_diam_inc(filter_soiltype = soil_type,
                 style = "band",
                 drought_bars = TRUE,
                 save_fig = FALSE, device = "pdf")
```

### Analysis Beech (Fagus)

#### Effect Size Over Time

```{r diameter-fagus-effect-size, fig.width=9.5, fig.height=5.8}
res_diameter_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "growth",
  resp_var = "diameter",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_diameter_fagus$temporal$plot)
```

#### SEM Without Interaction

```{r diameter-fagus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_diameter_fagus$sem$plot)
```

#### SEM With Interaction

```{r diameter-fagus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_diameter_fagus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "growth",
  resp_var = "diameter",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_diameter_fagus_int$sem$plot)
```

### Analysis Oak (Quercus)

#### Effect Size Over Time

```{r diameter-quercus-effect-size, fig.width=9.5, fig.height=5.8}
res_diameter_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "growth",
  resp_var = "diameter",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_diameter_quercus$temporal$plot)
```

#### SEM Without Interaction

```{r diameter-quercus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_diameter_quercus$sem$plot)
```

#### SEM With Interaction

```{r diameter-quercus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_diameter_quercus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "growth",
  resp_var = "diameter",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_diameter_quercus_int$sem$plot)
```

---

## Phenology: Leaf Development Stage

### Timeseries (Beech and Oak)

```{r phenology-ts, fig.width=9.5, fig.height=5.8}
plot_ts_phenology(style = "band",
                  filter_soiltype = soil_type,
                  save_fig = FALSE, device = "pdf")
```

### Analysis Beech (Fagus)

#### Effect Size Over Time

```{r phenology-fagus-effect-size, fig.width=9.5, fig.height=5.8}
res_phenology_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "phenology",
  resp_var = "stage",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_phenology_fagus$temporal$plot)
```

#### SEM Without Interaction

```{r phenology-fagus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_phenology_fagus$sem$plot)
```

#### SEM With Interaction

```{r phenology-fagus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_phenology_fagus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "phenology",
  resp_var = "stage",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_phenology_fagus_int$sem$plot)
```

### Analysis Oak (Quercus)

#### Effect Size Over Time

```{r phenology-quercus-effect-size, fig.width=9.5, fig.height=5.8}
res_phenology_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "phenology",
  resp_var = "stage",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_phenology_quercus$temporal$plot)
```

#### SEM Without Interaction

```{r phenology-quercus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_phenology_quercus$sem$plot)
```

#### SEM With Interaction

```{r phenology-quercus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_phenology_quercus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "phenology",
  resp_var = "stage",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_phenology_quercus_int$sem$plot)
```

---

## Quantum Yield (QY)

### Timeseries (Beech and Oak)

```{r qy-ts, fig.width=9.5, fig.height=5.8}
plot_ts_qy(filter_soiltype = soil_type,
           style = "band",
           drought_bars = TRUE,
           save_fig = FALSE, device = "pdf")
```

### Analysis Beech (Fagus)

#### Effect Size Over Time

```{r qy-fagus-effect-size, fig.width=9.5, fig.height=5.8}
res_qy_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "quantum_yield",
  resp_var = "qy",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_qy_fagus$temporal$plot)
```

#### SEM Without Interaction

```{r qy-fagus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_qy_fagus$sem$plot)
```

#### SEM With Interaction

```{r qy-fagus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_qy_fagus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "quantum_yield",
  resp_var = "qy",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_qy_fagus_int$sem$plot)
```

### Analysis Oak (Quercus)

#### Effect Size Over Time

```{r qy-quercus-effect-size, fig.width=9.5, fig.height=5.8}
res_qy_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "quantum_yield",
  resp_var = "qy",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_qy_quercus$temporal$plot)
```

#### SEM Without Interaction

```{r qy-quercus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_qy_quercus$sem$plot)
```

#### SEM With Interaction

```{r qy-quercus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_qy_quercus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "quantum_yield",
  resp_var = "qy",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_qy_quercus_int$sem$plot)
```

---

## Senescence: Percent Senesced

### Timeseries (Beech and Oak)

```{r senescence-pct-ts, fig.width=9.5, fig.height=5.8}
plot_ts_senescence_percent(filter_soiltype = soil_type,
                           style = "band",
                           drought_bars = FALSE,
                           save_fig = FALSE, device = "pdf")
```

### Analysis Beech (Fagus)

#### Effect Size Over Time

```{r senescence-pct-fagus-effect-size, fig.width=9.5, fig.height=5.8}
res_sen_pct_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "percent_senesced",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_sen_pct_fagus$temporal$plot)
```

#### SEM Without Interaction

```{r senescence-pct-fagus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_sen_pct_fagus$sem$plot)
```

#### SEM With Interaction

```{r senescence-pct-fagus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_sen_pct_fagus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "percent_senesced",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_sen_pct_fagus_int$sem$plot)
```

### Analysis Oak (Quercus)

#### Effect Size Over Time

```{r senescence-pct-quercus-effect-size, fig.width=9.5, fig.height=5.8}
res_sen_pct_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "percent_senesced",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_sen_pct_quercus$temporal$plot)
```

#### SEM Without Interaction

```{r senescence-pct-quercus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_sen_pct_quercus$sem$plot)
```

#### SEM With Interaction

```{r senescence-pct-quercus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_sen_pct_quercus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "percent_senesced",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_sen_pct_quercus_int$sem$plot)
```

---

## Senescence: Chlorophyll

### Timeseries (Beech and Oak)

```{r senescence-chl-ts, fig.width=9.5, fig.height=5.8}
plot_ts_senescence_chl(filter_soiltype = soil_type,
                       style = "band",
                       drought_bars = FALSE,
                       save_fig = FALSE, device = "pdf")
```

### Analysis Beech (Fagus)

#### Effect Size Over Time

```{r senescence-chl-fagus-effect-size, fig.width=9.5, fig.height=5.8}
res_sen_chl_fagus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "chlavg",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_sen_chl_fagus$temporal$plot)
```

#### SEM Without Interaction

```{r senescence-chl-fagus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_sen_chl_fagus$sem$plot)
```

#### SEM With Interaction

```{r senescence-chl-fagus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_sen_chl_fagus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "chlavg",
  species = "fagus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_sen_chl_fagus_int$sem$plot)
```

### Analysis Oak (Quercus)

#### Effect Size Over Time

```{r senescence-chl-quercus-effect-size, fig.width=9.5, fig.height=5.8}
res_sen_chl_quercus <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "chlavg",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = FALSE
)
print(res_sen_chl_quercus$temporal$plot)
```

#### SEM Without Interaction

```{r senescence-chl-quercus-sem-no-int, fig.width=9.5, fig.height=5.8}
print(res_sen_chl_quercus$sem$plot)
```

#### SEM With Interaction

```{r senescence-chl-quercus-sem-with-int, fig.width=9.5, fig.height=5.8}
res_sen_chl_quercus_int <- make_temporal_sem_combo(
  type = "tree",
  data_name = "senescence",
  resp_var = "chlavg",
  species = "quercus",
  soil_type = soil_type,
  include_interaction = TRUE
)
print(res_sen_chl_quercus_int$sem$plot)
```

---

# SWC Imputation: Treatment-Agnostic GAM

Soil water content (SWC) is measured manually at irregular intervals. For the SEMs above, SWC was sourced from the nearest observation within a ±7 day window. As a sensitivity check, we also build a **treatment-agnostic** daily SWC imputation using a GAM so that imputed values do not encode treatment information.

## GAM Model

$$\text{SWC}_{it} = f_1(t) + f_2(\text{SWP}_{\text{site},t}) + f_3(T_{\text{air},t}) + f_4(\text{VPD}_t) + f_5(\text{Rad}_t) + b_i + \varepsilon_{it}$$

where $f_1, \ldots, f_5$ are penalised thin-plate splines and $b_i$ is a random intercept per box.

### Data Preparation

```{r gam-meteo-daily}
meteo_file <- .resolve_path("data/raw/sensor_data/meteo_10min.dat")
meteo_raw <- read.table(
  meteo_file, sep = ",", header = FALSE, skip = 4,
  stringsAsFactors = FALSE,
  col.names = c("datetime", "id", "air_temp", "rh", "patm", "vpd", "radiation")
)

meteo_daily <- meteo_raw %>%
  mutate(datetime = ymd_hms(datetime), date = as.Date(datetime)) %>%
  mutate(across(c(air_temp, rh, vpd, radiation), as.numeric)) %>%
  group_by(date) %>%
  summarise(
    air_temp = mean(air_temp, na.rm = TRUE),
    rh       = mean(rh, na.rm = TRUE),
    vpd      = mean(vpd, na.rm = TRUE),
    radiation = mean(radiation, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r gam-swp-daily}
soil_file <- .resolve_path("data/raw/sensor_data/soil_10min.dat")
soil_raw <- read.table(
  soil_file, sep = ",", header = TRUE, skip = 2, stringsAsFactors = FALSE
) %>% slice(-1)

swp_site_daily <- soil_raw %>%
  mutate(datetime = ymd_hms(TS), date = as.Date(datetime)) %>%
  select(date, starts_with("kPa")) %>%
  mutate(across(-date, ~ as.numeric(na_if(.x, "NAN")))) %>%
  group_by(date) %>%
  summarise(across(everything(), \(x) mean(x, na.rm = TRUE)), .groups = "drop") %>%
  pivot_longer(starts_with("kPa"), names_to = "sensor_raw", values_to = "soil_mp") %>%
  filter(!is.na(soil_mp), soil_mp < 0) %>%
  mutate(swp_log10 = log10(-soil_mp)) %>%
  group_by(date) %>%
  summarise(swp_site = median(swp_log10, na.rm = TRUE), .groups = "drop")
```

```{r gam-panel-build}
swc_obs <- readr::read_csv(
  .resolve_path("data/interim/box_soilwater.csv"), show_col_types = FALSE
) %>%
  transmute(boxlabel = as.character(boxlabel), date = as.Date(date), swc_obs = as.numeric(swc))

date_min <- max(min(meteo_daily$date), min(swp_site_daily$date))
date_max <- min(max(meteo_daily$date), max(swp_site_daily$date))

panel_daily <- tidyr::expand_grid(
  swc_obs %>% distinct(boxlabel),
  tibble(date = seq.Date(date_min, date_max, by = "day"))
) %>%
  left_join(swc_obs, by = c("boxlabel", "date")) %>%
  left_join(meteo_daily, by = "date") %>%
  left_join(swp_site_daily, by = "date") %>%
  mutate(date_num = as.numeric(date), boxlabel = factor(boxlabel))

fit_data <- panel_daily %>%
  filter(!is.na(swc_obs), !is.na(swp_site), !is.na(air_temp), !is.na(vpd), !is.na(radiation))
```

### GAM Fit & Diagnostics

```{r gam-fit}
gam_swc <- mgcv::gam(
  swc_obs ~ s(date_num, k = 10) + s(swp_site, k = 8) +
    s(air_temp, k = 8) + s(vpd, k = 8) + s(radiation, k = 8) +
    s(boxlabel, bs = "re"),
  data = fit_data, method = "REML"
)
```

```{r gam-summary, results='markup'}
summary(gam_swc)
```

```{r gam-smooth-plots, fig.width=11, fig.height=8}
plot(gam_swc, pages = 1, shade = TRUE, residuals = TRUE, scale = 0)
```

### Prediction & Validation

```{r gam-predict}
pred_data <- panel_daily %>%
  filter(!is.na(swp_site), !is.na(air_temp), !is.na(vpd), !is.na(radiation))
pred <- predict(gam_swc, newdata = pred_data, se.fit = TRUE, type = "response")

gam_out <- pred_data %>%
  mutate(
    swc_hat    = pmin(pmax(as.numeric(pred$fit), 0), 100),
    swc_hat_se = as.numeric(pred$se.fit),
    swc_hat_lo = pmax(swc_hat - 1.96 * swc_hat_se, 0),
    swc_hat_hi = pmin(swc_hat + 1.96 * swc_hat_se, 100),
    swc_source = if_else(!is.na(swc_obs), "measured", "imputed_gam")
  ) %>%
  arrange(boxlabel, date)
```

```{r gam-metrics, results='markup'}
gam_out %>%
  filter(!is.na(swc_obs)) %>%
  summarise(n = n(),
            rmse = sqrt(mean((swc_obs - swc_hat)^2)),
            mae  = mean(abs(swc_obs - swc_hat)),
            r    = cor(swc_obs, swc_hat))
```

```{r gam-obs-vs-fit}
gam_out %>%
  filter(!is.na(swc_obs)) %>%
  ggplot(aes(swc_obs, swc_hat)) +
  geom_point(alpha = 0.35, size = 1.1, color = "grey25") +
  geom_abline(slope = 1, intercept = 0, color = "firebrick", linewidth = 0.7) +
  coord_equal() +
  labs(title = "Observed vs Fitted SWC", x = "Observed SWC", y = "Fitted SWC") +
  theme_linedraw()
```

```{r gam-residuals}
gam_res <- gam_out %>% filter(!is.na(swc_obs)) %>% mutate(resid = swc_obs - swc_hat)

p_res1 <- ggplot(gam_res, aes(swc_hat, resid)) +
  geom_point(alpha = 0.35, size = 1.0) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "firebrick") +
  labs(title = "Residuals vs Fitted", x = "Fitted SWC", y = "Residual") +
  theme_linedraw()

p_res2 <- ggplot(gam_res, aes(resid)) +
  geom_histogram(bins = 35, fill = "grey70", color = "white") +
  labs(title = "Residual Distribution", x = "Residual", y = "Count") +
  theme_linedraw()

p_res1
p_res2
```

### Example Box Trajectories

```{r gam-trajectories, fig.width=11, fig.height=7}
set.seed(42)
example_boxes <- gam_out %>% distinct(boxlabel) %>% slice_sample(n = 6) %>% pull(boxlabel)

gam_out %>%
  filter(boxlabel %in% example_boxes) %>%
  ggplot(aes(date, swc_hat)) +
  geom_ribbon(aes(ymin = swc_hat_lo, ymax = swc_hat_hi), fill = "grey80", alpha = 0.5) +
  geom_line(color = "steelblue", linewidth = 0.5) +
  geom_point(data = ~ dplyr::filter(.x, !is.na(swc_obs)), aes(y = swc_obs),
             color = "black", size = 1.1, alpha = 0.8) +
  facet_wrap(~ boxlabel, scales = "free_y", ncol = 3) +
  labs(title = "Daily SWC Trajectories (Lines = GAM, Points = Measured)", x = "Date", y = "SWC") +
  theme_linedraw()
```

### Export Imputed SWC

```{r gam-export}
readr::write_csv(gam_out, .resolve_path("data/interim/box_soilwater_daily_gam_agnostic.csv"))
```

---

# SWC Source Comparison: Influence on SEM Results

The SEMs above used measured SWC. Here we systematically compare SEM results when using imputed (GAM) SWC instead, to assess whether the choice of SWC source materially changes conclusions.

```{r swc-comp-controls}
swc_levels <- c("measured", "imputed_gam")
species_levels <- c("fagus", "quercus")
interaction_levels <- c(FALSE, TRUE)

swc_model_grid <- tribble(
  ~section, ~data_name, ~resp_var,
  "Chlorophyll",            "chlorophyll",  "chl",
  "Tree Condition",         "condition",    "condition",
  "Growth Height",          "growth",       "height",
  "Growth Diameter",        "growth",       "diameter",
  "Phenology Stage",        "phenology",    "stage",
  "Quantum Yield",          "quantum_yield","qy",
  "Senescence Percent",     "senescence",   "percent_senesced",
  "Senescence Chlorophyll", "senescence",   "chlavg"
)
```

```{r swc-comp-helpers}
fit_sem_swc <- function(data_name, resp_var, species, include_interaction, swc_source) {
  tryCatch(
    run_sem_for_trait(
      type = "tree", data_name = data_name, resp_var = resp_var,
      species = species, soil_type = soil_type,
      include_interaction = include_interaction,
      scale_all_numeric = TRUE, do_rfe = FALSE,
      swc_source = swc_source, force_run = FALSE
    ),
    error = function(e) { message("SEM fit failed: ", conditionMessage(e)); NULL }
  )
}

plot_sem_pair <- function(res_meas, res_imp, panel_title = NULL) {
  p_meas <- if (!is.null(res_meas) && !is.null(res_meas$plot)) {
    res_meas$plot + ggtitle("Measured SWC")
  } else {
    ggplot() + theme_void() + ggtitle("Measured SWC (missing)")
  }
  p_imp <- if (!is.null(res_imp) && !is.null(res_imp$plot)) {
    res_imp$plot + ggtitle("Imputed SWC")
  } else {
    ggplot() + theme_void() + ggtitle("Imputed SWC (missing)")
  }
  p <- p_meas | p_imp
  if (!is.null(panel_title)) p <- p + plot_annotation(title = panel_title)
  p
}

extract_sem_r2 <- function(res) {
  if (is.null(res) || is.null(res$sem)) {
    return(tibble(sem_r2_swc_marginal = NA_real_, sem_r2_swc_conditional = NA_real_,
                  sem_r2_resp_marginal = NA_real_, sem_r2_resp_conditional = NA_real_))
  }
  r2_tbl <- tryCatch(piecewiseSEM::rsquared(res$sem), error = function(e) NULL)
  if (is.null(r2_tbl)) {
    return(tibble(sem_r2_swc_marginal = NA_real_, sem_r2_swc_conditional = NA_real_,
                  sem_r2_resp_marginal = NA_real_, sem_r2_resp_conditional = NA_real_))
  }
  tibble(
    sem_r2_swc_marginal    = r2_tbl$Marginal[r2_tbl$Response == "swc"][1],
    sem_r2_swc_conditional = r2_tbl$Conditional[r2_tbl$Response == "swc"][1],
    sem_r2_resp_marginal   = r2_tbl$Marginal[r2_tbl$Response == "y"][1],
    sem_r2_resp_conditional = r2_tbl$Conditional[r2_tbl$Response == "y"][1]
  )
}

extract_sem_paths <- function(res) {
  empty <- tibble(factor = character(), effect_class = character(), path = character(),
                  estimate = double(), p_value = double(), abs_estimate = double())
  if (is.null(res)) return(empty)
  out <- list()
  if (!is.null(res$effects) && nrow(res$effects) > 0) {
    out[[length(out) + 1L]] <- bind_rows(
      res$effects %>% transmute(factor, effect_class = "main", path = "to_swc (a)", estimate = a, p_value = p_a),
      res$effects %>% transmute(factor, effect_class = "main", path = "to_target_direct (c')", estimate = c_direct, p_value = p_c),
      res$effects %>% transmute(factor, effect_class = "main", path = "to_target_indirect (a*b)", estimate = indirect, p_value = p_ind),
      res$effects %>% transmute(factor, effect_class = "main", path = "to_target_total", estimate = total, p_value = p_tot)
    ) %>% mutate(abs_estimate = abs(estimate))
  }
  if (!is.null(res$effects_int) && nrow(res$effects_int) > 0) {
    out[[length(out) + 1L]] <- bind_rows(
      res$effects_int %>% transmute(factor, effect_class = "interaction", path = "to_swc (a)", estimate = a, p_value = p_a),
      res$effects_int %>% transmute(factor, effect_class = "interaction", path = "to_target_direct (c')", estimate = c_direct, p_value = p_c),
      res$effects_int %>% transmute(factor, effect_class = "interaction", path = "to_target_indirect (a*b)", estimate = indirect, p_value = p_ind),
      res$effects_int %>% transmute(factor, effect_class = "interaction", path = "to_target_total", estimate = total, p_value = p_tot)
    ) %>% mutate(abs_estimate = abs(estimate))
  }
  if (!length(out)) return(empty)
  bind_rows(out)
}
```

## Fit SEM Grid (Measured vs Imputed)

```{r swc-comp-fit-all}
swc_fit_tasks <- tidyr::crossing(
  swc_model_grid,
  species = species_levels,
  include_interaction = interaction_levels,
  swc_source = swc_levels
) %>%
  mutate(result = purrr::pmap(
    list(data_name, resp_var, species, include_interaction, swc_source), fit_sem_swc
  ))

swc_metrics_long <- swc_fit_tasks %>%
  mutate(metrics = purrr::map(result, extract_sem_r2)) %>%
  tidyr::unnest(metrics) %>%
  mutate(target_label = paste0(section, " | ", species, if_else(include_interaction, " | int", " | no-int")))

swc_sem_paths_long <- swc_fit_tasks %>%
  mutate(sem_paths = purrr::map(result, extract_sem_paths)) %>%
  tidyr::unnest(sem_paths)
```

## Faceted 1:1 SEM Metrics

```{r swc-comp-1to1-metrics, fig.width=14, fig.height=10}
sem_metric_cols <- c("sem_r2_swc_marginal", "sem_r2_swc_conditional",
                     "sem_r2_resp_marginal", "sem_r2_resp_conditional")

swc_metrics_wide <- swc_metrics_long %>%
  select(section, data_name, resp_var, species, include_interaction, target_label,
         swc_source, all_of(sem_metric_cols)) %>%
  pivot_longer(cols = all_of(sem_metric_cols), names_to = "metric", values_to = "value") %>%
  pivot_wider(names_from = swc_source, values_from = value) %>%
  mutate(delta_imputed_minus_measured = imputed_gam - measured)

metrics_plot_df <- swc_metrics_wide %>%
  filter(is.finite(measured), is.finite(imputed_gam)) %>%
  mutate(
    metric = factor(metric,
      levels = sem_metric_cols,
      labels = c("SWC R2 Marginal", "SWC R2 Conditional",
                 "Target R2 Marginal", "Target R2 Conditional")),
    interaction_label = if_else(include_interaction, "With interaction", "No interaction")
  )

ggplot(metrics_plot_df, aes(x = measured, y = imputed_gam, color = species)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey40") +
  geom_point(size = 2.2, alpha = 0.9) +
  geom_text(aes(label = section), size = 2.8, check_overlap = TRUE, vjust = -0.6, show.legend = FALSE) +
  facet_grid(interaction_label ~ metric) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(title = "SEM performance by SWC source",
       subtitle = "Points above 1:1 line indicate higher metric with imputed SWC",
       x = "Measured SWC", y = "Imputed SWC", color = "Species") +
  theme_linedraw(base_size = 11) + theme(legend.position = "bottom")
```

## Factor Influence: Measured vs Imputed SWC

```{r swc-comp-factor-influence, fig.width=16, fig.height=10}
infl_df <- swc_sem_paths_long %>%
  filter(include_interaction == FALSE, effect_class == "main") %>%
  filter(path %in% c("to_swc (a)", "to_target_direct (c')", "to_target_indirect (a*b)", "to_target_total"))

summarise_importance <- function(df, species_keep = NULL) {
  x <- df
  if (!is.null(species_keep)) x <- x %>% filter(species == species_keep)
  x %>%
    group_by(swc_source, factor, path) %>%
    summarise(mean_abs = mean(abs_estimate, na.rm = TRUE),
              se_abs = sd(abs_estimate, na.rm = TRUE) / sqrt(dplyr::n()),
              .groups = "drop")
}

plot_importance_panel <- function(df_imp, title_txt) {
  ggplot(df_imp, aes(x = reorder(factor, mean_abs), y = mean_abs, fill = swc_source)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.72) +
    geom_errorbar(aes(ymin = pmax(0, mean_abs - se_abs), ymax = mean_abs + se_abs),
                  position = position_dodge(width = 0.8), width = 0.2) +
    facet_wrap(~ path, scales = "free_y", ncol = 2) + coord_flip() +
    labs(title = title_txt, x = "Factor", y = "Mean |effect|", fill = "SWC source") +
    theme_linedraw(base_size = 11) + theme(legend.position = "bottom")
}

p_all <- plot_importance_panel(summarise_importance(infl_df), "All species")
p_fag <- plot_importance_panel(summarise_importance(infl_df, "fagus"), "Fagus")
p_que <- plot_importance_panel(summarise_importance(infl_df, "quercus"), "Quercus")

(p_all | p_fag | p_que) +
  patchwork::plot_layout(guides = "collect") +
  patchwork::plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom", plot.tag = element_text(face = "bold"))
```

## Full SEM Figure Comparison

```{r swc-comp-sem-figures, results='asis', fig.width=16, fig.height=7}
get_swc_fit <- function(sec, dn, rv, sp, int, src) {
  swc_fit_tasks %>%
    filter(.data$section == sec, .data$data_name == dn, .data$resp_var == rv,
           .data$species == sp, .data$include_interaction == int, .data$swc_source == src) %>%
    pull(result) %>% .[[1]]
}

for (i in seq_len(nrow(swc_model_grid))) {
  row <- swc_model_grid[i, ]
  cat("\n\n### ", row$section, "\n", sep = "")
  for (sp in species_levels) {
    cat("\n\n#### Species: ", sp, "\n", sep = "")

    res_meas <- get_swc_fit(row$section, row$data_name, row$resp_var, sp, FALSE, "measured")
    res_imp  <- get_swc_fit(row$section, row$data_name, row$resp_var, sp, FALSE, "imputed_gam")
    cat("\n\n##### No Interaction\n")
    print(plot_sem_pair(res_meas, res_imp,
      paste0("SEM no interaction: ", row$data_name, "/", row$resp_var, "/", sp)))

    res_meas_int <- get_swc_fit(row$section, row$data_name, row$resp_var, sp, TRUE, "measured")
    res_imp_int  <- get_swc_fit(row$section, row$data_name, row$resp_var, sp, TRUE, "imputed_gam")
    cat("\n\n##### With Interaction\n")
    print(plot_sem_pair(res_meas_int, res_imp_int,
      paste0("SEM with interaction: ", row$data_name, "/", row$resp_var, "/", sp)))
  }
}
```

---

# Session Info

```{r session-info}
sessionInfo()
```
