---
title: "ALINV - SEM Aggregation Matrices"
output:
  html_document:
    toc: true
    toc_depth: 2
params:
  scenario_label: "both soils (with soil as treatment)"
  soil_filter: "both"
  include_soil_treatment: true
  analysis_date: !r as.character(Sys.Date())
  output_root: "output"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 5.8,
  dpi = 120
)

source_path_candidates <- c(
  "functions/_source.R",
  "../functions/_source.R",
  "./functions/_source.R"
)
source_path <- source_path_candidates[file.exists(source_path_candidates)][1]
if (is.na(source_path)) {
  stop("Could not locate functions/_source.R for notebook setup.")
}
source(source_path)

source_dir <- dirname(source_path)
source(file.path(source_dir, "3-effect-size-factorial.R"))
source(file.path(source_dir, "4-structural-equation-model.R"))

library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)

analysis_ctx <- alinv_init_notebook_context(if (exists("params")) params else NULL)
scenario_label <- analysis_ctx$scenario_label
soil_type <- analysis_ctx$soil_filter
include_soil_treatment <- analysis_ctx$include_soil_treatment
analysis_root <- analysis_ctx$analysis_root
analysis_data_root <- analysis_ctx$analysis_data_root
heatmap_export_dir <- alinv_data_path("structural_equation_models", "heatmap_data", create_dir = TRUE)
```

# Overview

This notebook focuses only on the aggregated SEM heatmaps across tree response variables.

Current render scenario: `r scenario_label`. Soil filtering is set to `r soil_type`, and soil is `r if (include_soil_treatment) "included as an explicit treatment contrast." else "pooled across both soils and omitted as a treatment contrast."`

Phase-wise SEM panels were removed because per-phase subsets do not contain enough data for stable SEM fitting.

The underlying heatmap matrices are exported as CSV files under `r heatmap_export_dir`. These CSVs can be replotted later with `scripts/6-plot-sem-heatmaps-from-csv.R`.

Interpretation of effects in the mediation SEM:

- Direct effect: treatment -> target (path c').
- Indirect effect: treatment -> SWC (path a) multiplied by SWC -> target (path b), i.e. a x b.
- Combined effect (total): direct + indirect.

Important clarification: the treatment -> SWC path alone is not the full indirect effect. It is one component (path a) used to compute the indirect effect together with path b.

```{r config}
species_vec <- c("fagus", "quercus")
include_interaction <- FALSE

resp_labels <- c(
  chl = "Chlorophyll",
  condition = "Condition",
  height = "Height",
  diameter = "Diameter",
  height_inc_t0_rel = "Height rel. inc t0",
  diameter_inc_t0_rel = "Diameter rel. inc t0",
  height_inc_phase_rel = "Height rel. inc phase",
  diameter_inc_phase_rel = "Diameter rel. inc phase",
  qy = "Quantum yield",
  percent_senesced = "Senescence (%)",
  chlavg = "Senescence chlorophyll",
  stage = "Phenology stage"
)

treat_labels <- sem_heatmap_treatment_labels()
heatmap_specs <- sem_heatmap_specs()
```

```{r sem-collect}
safe_run_sem <- function(...) {
  tryCatch(run_sem_for_trait(...), error = function(e) {
    message("Skipping SEM fit: ", conditionMessage(e))
    NULL
  })
}

collect_sem_matrix_data <- function(species, include_interaction = FALSE) {
  tree_grid <- get_data_var_grid() %>%
    dplyr::filter(type == "tree")

  out <- vector("list", nrow(tree_grid))

  for (k in seq_len(nrow(tree_grid))) {
    g <- tree_grid[k, ]

    sem_res <- safe_run_sem(
      type = g$type,
      data_name = g$data,
      resp_var = g$resp_var,
      species = species,
      soil_type = soil_type,
      include_soil_treatment = include_soil_treatment,
      include_interaction = include_interaction,
      phase_window = "all",
      scale_all_numeric = TRUE,
      swc_source = "measured",
      force_run = FALSE
    )

    if (is.null(sem_res)) next

    if (is.null(sem_res$matrix_data) && !is.null(sem_res$effects) && nrow(sem_res$effects)) {
      sem_res$matrix_data <- build_sem_matrix_data(
        effects_main = sem_res$effects,
        effects_int = sem_res$effects_int,
        resp_var = g$resp_var,
        species = species,
        include_interaction = include_interaction,
        swc_source = "measured",
        phase = "all"
      )
    }

    if (is.null(sem_res) || is.null(sem_res$matrix_data)) next

    out[[k]] <- sem_res$matrix_data %>%
      dplyr::mutate(
        data_name = g$data,
        resp_var_raw = g$resp_var
      )
  }

  dplyr::bind_rows(out)
}

sem_all_species <- purrr::map_dfr(
  species_vec,
  ~ collect_sem_matrix_data(species = .x, include_interaction = include_interaction)
)

if (!nrow(sem_all_species)) {
  sem_all_species <- tibble::tibble(
    treatment = character(),
    response_var = character(),
    path_type = character(),
    estimate = double(),
    p_value = double(),
    effect_class = character(),
    significant = logical(),
    estimate_sig = double(),
    species = character(),
    include_interaction = logical(),
    swc_source = character(),
    phase = character(),
    data_name = character(),
    resp_var_raw = character()
  )
}

shared_heatmap_limit <- compute_shared_sem_heatmap_limit(sem_all_species)

purrr::walk(
  species_vec,
  ~ write_sem_heatmap_csvs(
    matrix_df = sem_all_species,
    species = .x,
    export_dir = heatmap_export_dir,
    resp_labels = resp_labels,
    treat_labels = treat_labels
  )
)
```

# Heatmaps Per Species

Shared color-scale limit across all heatmaps in this scenario: `r round(shared_heatmap_limit, 3)`.

```{r sem-matrix-species-loop, fig.width=12, fig.height=6}
plot_sem_heatmaps_for_species <- function(species_name) {
  species_df <- sem_all_species %>%
    dplyr::filter(.data$species == species_name, .data$effect_class == "main")

  if (!nrow(species_df)) {
    return(list(
      ggplot() + theme_void() + ggtitle(paste("No SEM matrix data available for", species_name))
    ))
  }

  purrr::map(seq_len(nrow(heatmap_specs)), function(i) {
    spec <- heatmap_specs[i, ]
    panel_df <- prepare_sem_heatmap_panel(
      matrix_df = species_df,
      path_type = spec$path_type,
      resp_labels = resp_labels,
      treat_labels = treat_labels
    )

    plot_sem_heatmap_panel(
      panel_df = panel_df,
      limit = shared_heatmap_limit,
      title = paste(species_name, "-", spec$panel_title)
    )
  })
}

for (sp in species_vec) {
  cat("\n\n## Species:", sp, "\n")

  plot_list <- plot_sem_heatmaps_for_species(sp)
  for (p in plot_list) {
    print(p)
  }
}
```

# Session Info

```{r session-info}
sessionInfo()
```
