---
title: "ALINV - SEM Aggregation Matrices"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 5.8,
  dpi = 120
)

source_path_candidates <- c(
  "functions/_source.R",
  "../functions/_source.R",
  "./functions/_source.R"
)
source_path <- source_path_candidates[file.exists(source_path_candidates)][1]
if (is.na(source_path)) {
  stop("Could not locate functions/_source.R for notebook setup.")
}
source(source_path)

source_dir <- dirname(source_path)
source(file.path(source_dir, "3-effect-size-factorial.R"))
source(file.path(source_dir, "4-structural-equation-model.R"))

library(dplyr)
library(tidyr)
library(ggplot2)
```

# Overview

This notebook focuses only on SEM aggregation outputs.

Interpretation of effects in the mediation SEM:

- Direct effect: treatment -> target (path c').
- Indirect effect: treatment -> SWC (path a) multiplied by SWC -> target (path b), i.e. a x b.
- Combined effect (total): direct + indirect.

Important clarification: the treatment -> SWC path alone is not the full indirect effect. It is one component (path a) used to compute the indirect effect together with path b.

```{r config}
soil_type <- "both"
species_vec <- c("fagus", "quercus")
include_interaction <- FALSE
p_sig <- 0.05
phase_windows <- c("until June", "July-August", "September-end", "all")

resp_labels <- c(
  chl = "Chlorophyll",
  condition = "Condition",
  height = "Height",
  diameter = "Diameter",
  height_inc_t0_rel = "Height rel. inc t0",
  diameter_inc_t0_rel = "Diameter rel. inc t0",
  height_inc_phase_rel = "Height rel. inc phase",
  diameter_inc_phase_rel = "Diameter rel. inc phase",
  qy = "Quantum yield",
  percent_senesced = "Senescence (%)",
  chlavg = "Senescence chlorophyll",
  doy = "Phenology (DOY)"
)

treat_labels <- c(
  "robiniawith-robinia" = "Robinia: without -> with",
  "precipitationdrought" = "Precipitation: control -> drought",
  "culturemixed" = "Culture: mono -> mixed",
  "soiltypeinoc-robinia" = "Soil: beech -> robinia",
  "extreme_eventTRUE" = "Extreme event: no -> yes"
)
```

```{r sem-collect}
safe_run_sem <- function(...) {
  tryCatch(run_sem_for_trait(...), error = function(e) {
    message("Skipping SEM fit: ", conditionMessage(e))
    NULL
  })
}

collect_sem_matrix_data <- function(species, include_interaction = FALSE) {
  tree_grid <- get_data_var_grid() %>%
    dplyr::filter(type == "tree")

  out <- vector("list", nrow(tree_grid))

  for (k in seq_len(nrow(tree_grid))) {
    g <- tree_grid[k, ]

    sem_res <- safe_run_sem(
      type = g$type,
      data_name = g$data,
      resp_var = g$resp_var,
      species = species,
      soil_type = soil_type,
      include_interaction = include_interaction,
      phase_window = "all",
      scale_all_numeric = TRUE,
      swc_source = "measured",
      force_run = FALSE
    )

    if (is.null(sem_res)) next

    # Backward compatibility: older cached SEM objects may not contain matrix_data.
    if (is.null(sem_res$matrix_data) && !is.null(sem_res$effects) && nrow(sem_res$effects)) {
      sem_res$matrix_data <- build_sem_matrix_data(
        effects_main = sem_res$effects,
        effects_int = sem_res$effects_int,
        resp_var = g$resp_var,
        species = species,
        include_interaction = include_interaction,
        swc_source = "measured",
        phase = "all"
      )
    }

    # If cached object is too old/incomplete, force a refresh once.
    if (is.null(sem_res$matrix_data)) {
      sem_res <- safe_run_sem(
        type = g$type,
        data_name = g$data,
        resp_var = g$resp_var,
        species = species,
        soil_type = soil_type,
        include_interaction = include_interaction,
        phase_window = "all",
        scale_all_numeric = TRUE,
        swc_source = "measured",
        force_run = TRUE
      )
    }

    if (is.null(sem_res) || is.null(sem_res$matrix_data)) next

    out[[k]] <- sem_res$matrix_data %>%
      dplyr::mutate(
        data_name = g$data,
        resp_var_raw = g$resp_var
      )
  }

  dplyr::bind_rows(out)
}

collect_phase_sem_effects <- function(species, include_interaction = FALSE) {
  tree_grid <- get_data_var_grid() %>%
    dplyr::filter(type == "tree")

  out <- list()
  idx <- 1L

  for (ph in phase_windows) {
    for (k in seq_len(nrow(tree_grid))) {
      g <- tree_grid[k, ]

      sem_res <- safe_run_sem(
        type = g$type,
        data_name = g$data,
        resp_var = g$resp_var,
        species = species,
        soil_type = soil_type,
        include_interaction = include_interaction,
        phase_window = ph,
        scale_all_numeric = TRUE,
        swc_source = "measured",
        force_run = FALSE
      )

      if (is.null(sem_res) || is.null(sem_res$effects) || !nrow(sem_res$effects)) next

      eff <- sem_res$effects

      out[[idx]] <- eff %>%
        dplyr::mutate(
          phase = ph,
          species = species,
          data_name = g$data,
          target = g$resp_var
        )
      idx <- idx + 1L
    }
  }

  dplyr::bind_rows(out)
}

summarize_phase_sem_effects <- function(eff_df) {
  if (!nrow(eff_df)) return(tibble::tibble())

  a_paths <- eff_df %>%
    dplyr::group_by(.data$species, .data$phase, .data$factor) %>%
    dplyr::summarise(
      estimate = mean(.data$a, na.rm = TRUE),
      n_targets = dplyr::n_distinct(.data$target),
      path_type = "treatment_to_swc",
      .groups = "drop"
    ) %>%
    dplyr::rename(from = .data$factor)

  c_paths <- eff_df %>%
    dplyr::group_by(.data$species, .data$phase, .data$factor) %>%
    dplyr::summarise(
      estimate = mean(.data$c_direct, na.rm = TRUE),
      n_targets = dplyr::n_distinct(.data$target),
      path_type = "treatment_to_target",
      .groups = "drop"
    ) %>%
    dplyr::rename(from = .data$factor)

  b_paths <- eff_df %>%
    dplyr::group_by(.data$species, .data$phase, .data$target) %>%
    dplyr::summarise(b = dplyr::first(.data$b), .groups = "drop") %>%
    dplyr::group_by(.data$species, .data$phase) %>%
    dplyr::summarise(
      estimate = mean(.data$b, na.rm = TRUE),
      n_targets = dplyr::n(),
      path_type = "swc_to_target",
      .groups = "drop"
    ) %>%
    dplyr::mutate(from = "swc")

  dplyr::bind_rows(a_paths, c_paths, b_paths)
}

plot_phase_sem_arrow <- function(sum_df, species_name, phase_name) {
  df_phase <- sum_df %>%
    dplyr::filter(.data$species == species_name, .data$phase == phase_name)

  if (!nrow(df_phase)) {
    return(ggplot() + theme_void() + ggtitle(paste("No phase SEM summary for", species_name, "-", phase_name)))
  }

  y_map <- c(
    "extreme_eventTRUE" = 5,
    "robiniawith-robinia" = 4,
    "precipitationdrought" = 3,
    "soiltypeinoc-robinia" = 2,
    "culturemixed" = 1
  )

  nodes <- tibble::tibble(
    node = c(names(y_map), "swc", "target_mean"),
    x = c(rep(0.8, length(y_map)), 3.3, 5.8),
    y = c(unname(y_map), 3, 3)
  )

  edges <- df_phase %>%
    dplyr::mutate(
      to = dplyr::case_when(
        .data$path_type == "treatment_to_swc" ~ "swc",
        .data$path_type == "treatment_to_target" ~ "target_mean",
        .data$path_type == "swc_to_target" ~ "target_mean",
        TRUE ~ NA_character_
      )
    ) %>%
    dplyr::filter(!is.na(.data$to)) %>%
    dplyr::left_join(nodes, by = c("from" = "node")) %>%
    dplyr::rename(x_from = .data$x, y_from = .data$y) %>%
    dplyr::left_join(nodes, by = c("to" = "node")) %>%
    dplyr::rename(x_to = .data$x, y_to = .data$y) %>%
    dplyr::mutate(
      col_dir = dplyr::if_else(.data$estimate >= 0, "positive", "negative"),
      edge_size = scales::rescale(abs(.data$estimate), to = c(0.35, 2.2), from = range(abs(.data$estimate), na.rm = TRUE)),
      lbl = sprintf("%.2f", .data$estimate)
    )

  ggplot() +
    geom_segment(
      data = edges,
      aes(
        x = .data$x_from,
        y = .data$y_from,
        xend = .data$x_to,
        yend = .data$y_to,
        size = .data$edge_size,
        colour = .data$col_dir
      ),
      arrow = arrow(length = unit(0.24, "cm"), type = "open"),
      lineend = "round",
      alpha = 0.9
    ) +
    geom_text(
      data = edges,
      aes(
        x = (.data$x_from + .data$x_to) / 2,
        y = (.data$y_from + .data$y_to) / 2 + 0.14,
        label = .data$lbl
      ),
      size = 3
    ) +
    geom_label(
      data = nodes,
      aes(
        x = .data$x,
        y = .data$y,
        label = dplyr::recode(.data$node,
          !!!treat_labels,
          swc = "SWC",
          target_mean = "Target (mean)",
          .default = .data$node
        )
      ),
      size = 3.1,
      label.size = 0.22,
      fill = "white",
      label.padding = unit(0.12, "lines")
    ) +
    scale_colour_manual(values = c(positive = "steelblue4", negative = "indianred3"), name = "Direction") +
    scale_size_identity() +
    coord_cartesian(xlim = c(0.2, 6.6), ylim = c(0.3, 5.7), expand = FALSE) +
    labs(
      title = paste0(
        species_name,
        " - phase: ",
        phase_name,
        " (mean coefficients across targets)"
      )
    ) +
    theme_minimal(base_size = 11) +
    theme(
      axis.text = element_blank(),
      axis.title = element_blank(),
      panel.grid = element_blank(),
      legend.position = "bottom"
    )
}
```

# Phase-Wise Mean SEM Arrows

This section fits one SEM per target and per phase, then averages path coefficients across targets within each phase. The `all` phase panel is computed from SEM fits on all data per target (not by averaging phase subsets).

```{r phase-wise-sem-arrows, fig.width=12, fig.height=6}
for (sp in species_vec) {
  cat("\n\n## Species:", sp, "\n")

  phase_eff <- collect_phase_sem_effects(species = sp, include_interaction = include_interaction)

  if (!nrow(phase_eff)) {
    cat("No phase-wise SEM effects available for", sp, "\n")
    next
  }

  phase_sum <- summarize_phase_sem_effects(phase_eff)

  for (ph in phase_windows) {
    print(plot_phase_sem_arrow(phase_sum, species_name = sp, phase_name = ph))
  }
}
```

```{r sem-plot-helpers}
plot_treatment_target_matrix <- function(df_all, effect_type = c("direct", "indirect", "total"), species_name) {
  effect_type <- match.arg(effect_type)

  df_plot <- df_all %>%
    dplyr::filter(
      .data$species == species_name,
      .data$effect_class == "main",
      .data$path_type == effect_type,
      .data$treatment != "swc"
    ) %>%
    dplyr::mutate(
      target_label = dplyr::recode(.data$response_var, !!!resp_labels, .default = .data$response_var),
      treatment_label = dplyr::recode(.data$treatment, !!!treat_labels, .default = .data$treatment),
      est_sig = dplyr::if_else(!is.na(.data$p_value) & .data$p_value < p_sig, .data$estimate, NA_real_),
      txt = dplyr::if_else(is.na(.data$est_sig), "", sprintf("%.2f", .data$est_sig))
    )

  if (!nrow(df_plot)) {
    return(ggplot() + theme_void() + ggtitle(paste("No", effect_type, "effects available for", species_name)))
  }

  max_abs <- max(abs(df_plot$estimate), na.rm = TRUE)
  if (!is.finite(max_abs) || max_abs == 0) max_abs <- 1

  ggplot(df_plot, aes(x = target_label, y = treatment_label)) +
    geom_tile(aes(fill = est_sig), color = "grey85", linewidth = 0.3) +
    geom_text(aes(label = txt), size = 3) +
    scale_fill_gradient2(
      low = "indianred3",
      mid = "white",
      high = "steelblue4",
      midpoint = 0,
      limits = c(-max_abs, max_abs),
      na.value = "white",
      name = "Std. effect"
    ) +
    labs(
      title = paste0(species_name, " - ", toupper(substr(effect_type, 1, 1)), substring(effect_type, 2), " effects"),
      x = "Target variables",
      y = "Treatment"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid = element_blank(),
      axis.text.x = element_text(angle = 35, hjust = 1),
      legend.position = "right"
    )
}

plot_swc_target_panel <- function(df_all, species_name) {
  df_plot <- df_all %>%
    dplyr::filter(
      .data$species == species_name,
      .data$effect_class == "main",
      .data$path_type == "swc_to_target"
    ) %>%
    dplyr::mutate(
      target_label = dplyr::recode(.data$response_var, !!!resp_labels, .default = .data$response_var),
      row_label = "SWC -> target",
      est_sig = dplyr::if_else(!is.na(.data$p_value) & .data$p_value < p_sig, .data$estimate, NA_real_),
      txt = dplyr::if_else(is.na(.data$est_sig), "", sprintf("%.2f", .data$est_sig))
    )

  if (!nrow(df_plot)) {
    return(ggplot() + theme_void() + ggtitle(paste("No SWC -> target effects available for", species_name)))
  }

  max_abs <- max(abs(df_plot$estimate), na.rm = TRUE)
  if (!is.finite(max_abs) || max_abs == 0) max_abs <- 1

  ggplot(df_plot, aes(x = target_label, y = row_label)) +
    geom_tile(aes(fill = est_sig), color = "grey85", linewidth = 0.3) +
    geom_text(aes(label = txt), size = 3) +
    scale_fill_gradient2(
      low = "indianred3",
      mid = "white",
      high = "steelblue4",
      midpoint = 0,
      limits = c(-max_abs, max_abs),
      na.value = "white",
      name = "Std. effect"
    ) +
    labs(
      title = paste0(species_name, " - SWC -> target"),
      x = "Target variables",
      y = NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid = element_blank(),
      axis.text.x = element_text(angle = 35, hjust = 1),
      legend.position = "right"
    )
}

plot_treatment_swc_panel <- function(df_all, species_name) {
  df_plot <- df_all %>%
    dplyr::filter(
      .data$species == species_name,
      .data$effect_class == "main",
      .data$path_type == "treatment_to_swc",
      .data$treatment != "swc"
    ) %>%
    dplyr::distinct(.data$treatment, .keep_all = TRUE) %>%
    dplyr::mutate(
      treatment_label = dplyr::recode(.data$treatment, !!!treat_labels, .default = .data$treatment),
      col_label = "SWC",
      est_sig = dplyr::if_else(!is.na(.data$p_value) & .data$p_value < p_sig, .data$estimate, NA_real_),
      txt = dplyr::if_else(is.na(.data$est_sig), "", sprintf("%.2f", .data$est_sig))
    )

  if (!nrow(df_plot)) {
    return(ggplot() + theme_void() + ggtitle(paste("No treatment -> SWC effects available for", species_name)))
  }

  max_abs <- max(abs(df_plot$estimate), na.rm = TRUE)
  if (!is.finite(max_abs) || max_abs == 0) max_abs <- 1

  ggplot(df_plot, aes(x = col_label, y = treatment_label)) +
    geom_tile(aes(fill = est_sig), color = "grey85", linewidth = 0.3) +
    geom_text(aes(label = txt), size = 3) +
    scale_fill_gradient2(
      low = "indianred3",
      mid = "white",
      high = "steelblue4",
      midpoint = 0,
      limits = c(-max_abs, max_abs),
      na.value = "white",
      name = "Std. effect"
    ) +
    labs(
      title = paste0(species_name, " - treatment -> SWC (path a)"),
      x = NULL,
      y = "Treatment"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid = element_blank(),
      legend.position = "right"
    )
}
```

# Matrices Per Species

```{r sem-matrix-species-loop, fig.width=12, fig.height=6}
for (sp in species_vec) {
  cat("\n\n## Species:", sp, "\n")

  sem_all <- collect_sem_matrix_data(species = sp, include_interaction = include_interaction)

  if (!nrow(sem_all)) {
    cat("No SEM matrix data available for", sp, "\n")
    next
  }

  print(plot_treatment_target_matrix(sem_all, effect_type = "direct", species_name = sp))
  print(plot_treatment_target_matrix(sem_all, effect_type = "indirect", species_name = sp))
  print(plot_treatment_target_matrix(sem_all, effect_type = "total", species_name = sp))

  print(plot_treatment_swc_panel(sem_all, species_name = sp))
  print(plot_swc_target_panel(sem_all, species_name = sp))
}
```

# Session Info

```{r session-info}
sessionInfo()
```