---
title: "ALINV - Data QC"
date: "`r format(Sys.Date(), '%d %b %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    highlight: espresso
    theme: journal
params:
  scenario_label: "both soils (with soil as treatment)"
  soil_filter: "both"
  include_soil_treatment: true
  analysis_date: !r as.character(Sys.Date())
  output_root: "output"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

# Resolve project root robustly from the notebook render context
source_path_candidates <- c(
  "functions/_source.R",
  "../functions/_source.R",
  "./functions/_source.R"
)
source_path <- source_path_candidates[file.exists(source_path_candidates)][1]
if (is.na(source_path)) {
  stop("Could not locate functions/_source.R for notebook setup.")
}
source(source_path)

source_dir <- dirname(source_path)
proj_root <- normalizePath(file.path(source_dir, ".."), winslash = "/", mustWork = TRUE)

library(dplyr)
library(tidyr)
library(knitr)

analysis_ctx <- alinv_init_notebook_context(if (exists("params")) params else NULL)
scenario_label <- analysis_ctx$scenario_label
soil_type <- analysis_ctx$soil_filter
include_soil_treatment <- analysis_ctx$include_soil_treatment
analysis_root <- analysis_ctx$analysis_root
analysis_data_root <- analysis_ctx$analysis_data_root
```

# Overview

This QC notebook summarises the cleaned input data used for the current scenario render.

- Scenario: `r scenario_label`
- Soil filter: `r soil_type`
- Soil modeled as treatment: `r include_soil_treatment`
- Scenario output root: `r analysis_root`

# Phenology Cleaning

The phenology cleaner uses only the raw `Stage1`–`Stage4` dates from the workbook.

- `DOY_S1`–`DOY_S4` are ignored as source data and recomputed from the cleaned stage dates.
- Missing intermediate stages remain missing.
- If an earlier stage was recorded later than a later stage, the later-stage earlier date is kept and the conflicting earlier-stage date is nulled.
- The cleaned transition table is then used to rebuild the long phenology time series for the existing temporal models.

```{r phenology-load}
pheno_transitions <- get_data("tree", "phenology_transitions") %>%
  alinv_apply_soil_context(
    soil_filter = soil_type,
    include_soil_treatment = include_soil_treatment
  )

pheno_long <- get_data("tree", "phenology") %>%
  alinv_apply_soil_context(
    soil_filter = soil_type,
    include_soil_treatment = include_soil_treatment
  )
```

## Cleaning Summary

```{r phenology-qc-summary}
pheno_summary <- pheno_transitions %>%
  summarise(
    stage_rows = n(),
    retained_dates = sum(stage_date_source == "retained", na.rm = TRUE),
    raw_missing_dates = sum(stage_date_source == "missing_raw", na.rm = TRUE),
    stage_order_conflicts_nulled = sum(stage_order_conflict_nulled, na.rm = TRUE),
    affected_tree_count = n_distinct(tree_id[stage_order_conflict_nulled])
  )

kable(pheno_summary, caption = "Phenology cleaning summary")
```

## Affected Tree IDs

```{r phenology-qc-affected}
affected_trees <- pheno_transitions %>%
  filter(stage_order_conflict_nulled) %>%
  distinct(tree_id, species, soiltype, robinia, precipitation, culture) %>%
  arrange(species, tree_id)

if (nrow(affected_trees)) {
  kable(affected_trees, caption = "Trees with stage-order conflicts that were nulled during cleaning")
} else {
  cat("No stage-order conflicts were detected in this scenario subset.")
}
```

## Stage-Level Details

```{r phenology-qc-details}
pheno_detail <- pheno_transitions %>%
  filter(stage_order_conflict_nulled | stage_date_source == "missing_raw") %>%
  select(tree_id, species, stage, stage_date_raw, stage_date, stage_date_source) %>%
  arrange(species, tree_id, stage)

if (nrow(pheno_detail)) {
  kable(pheno_detail, caption = "Phenology rows that remained missing or were nulled during cleaning")
} else {
  cat("No missing or nulled phenology rows to report.")
}
```

## Monotonicity Checks

```{r phenology-qc-monotonic}
transition_monotonic <- pheno_transitions %>%
  filter(!is.na(stage_date)) %>%
  arrange(tree_id, stage) %>%
  group_by(tree_id) %>%
  summarise(
    is_monotonic = all(diff(stage_date) >= 0),
    .groups = "drop"
  )

long_monotonic <- pheno_long %>%
  arrange(tree_id, date) %>%
  group_by(tree_id) %>%
  summarise(
    is_monotonic = all(diff(stage) >= 0),
    .groups = "drop"
  )

mono_tbl <- tibble(
  check = c("Transition-stage dates", "Long stage-over-time table"),
  non_monotonic_trees = c(
    sum(!transition_monotonic$is_monotonic, na.rm = TRUE),
    sum(!long_monotonic$is_monotonic, na.rm = TRUE)
  )
)

kable(mono_tbl, caption = "Phenology monotonicity checks")
```

# Dataset Counts

## Tree and Box Counts

```{r scenario-counts}
tree_meta_scenario <- get_meta("tree") %>%
  alinv_apply_soil_context(
    soil_filter = soil_type,
    include_soil_treatment = include_soil_treatment
  )

box_meta_scenario <- get_meta("box") %>%
  alinv_apply_soil_context(
    soil_filter = soil_type,
    include_soil_treatment = include_soil_treatment
  )

tree_counts <- tree_meta_scenario %>%
  count(species, soiltype, robinia, precipitation, culture, name = "n_trees") %>%
  arrange(species, soiltype, robinia, precipitation, culture)

box_counts <- box_meta_scenario %>%
  count(soiltype, robinia, precipitation, culture, name = "n_boxes") %>%
  arrange(soiltype, robinia, precipitation, culture)

kable(tree_counts, caption = "Scenario-specific tree counts by treatment combination")
kable(box_counts, caption = "Scenario-specific box counts by treatment combination")
```

## Measurement Row Counts

```{r dataset-row-counts}
dataset_row_counts <- tibble::tribble(
  ~dataset, ~n_rows,
  "tree_growth", nrow(get_data("tree", "growth") %>% alinv_filter_by_soil(soil_filter = soil_type)),
  "tree_chlorophyll", nrow(get_data("tree", "chlorophyll") %>% alinv_filter_by_soil(soil_filter = soil_type)),
  "tree_condition", nrow(get_data("tree", "condition") %>% alinv_filter_by_soil(soil_filter = soil_type)),
  "tree_phenology", nrow(pheno_long),
  "tree_phenology_transitions", nrow(pheno_transitions),
  "tree_quantum_yield", nrow(get_data("tree", "quantum_yield") %>% alinv_filter_by_soil(soil_filter = soil_type)),
  "tree_senescence", nrow(get_data("tree", "senescence") %>% alinv_filter_by_soil(soil_filter = soil_type)),
  "box_soilwater", nrow(get_data("box", "soilwater") %>% alinv_filter_by_soil(soil_filter = soil_type)),
  "box_respiration", nrow(get_data("box", "respiration") %>% alinv_filter_by_soil(soil_filter = soil_type))
)

kable(dataset_row_counts, caption = "Scenario-specific row counts for key datasets")
```

# Session Info

```{r session-info}
sessionInfo()
```
