---
title: "ALINV - Agnostic SWC Imputation with GAM"
date: "`r format(Sys.Date(), '%d %b %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    theme: journal
    highlight: espresso
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 5
)

# Resolve project root whether knitting from repo root or notebooks/.
root_candidates <- c(getwd(), normalizePath(file.path(getwd(), ".."), mustWork = FALSE))
proj_root <- root_candidates[dir.exists(file.path(root_candidates, "data", "interim"))][1]
if (is.na(proj_root) || !nzchar(proj_root)) {
  stop("Could not locate project root containing data/interim.", call. = FALSE)
}
setwd(proj_root)

source("functions/_source.R")

library(tidyverse)
library(lubridate)
library(mgcv)
library(scales)
```

# Goal

Build a daily SWC time series for all boxes using an **agnostic** model:

- Use measured `swc` (irregular, all boxes)
- Use measured SWP signal from sensors (`kPa*`)
- Use site-level daily meteo (`air_temp`, `rh`, `vpd`, `radiation`)
- Avoid treatment predictors (`soiltype`, `culture`, `robinia`, `precipitation`)

# Data Wrangling

## Meteo to daily

```{r meteo-daily}
meteo_file <- .resolve_path("data/raw/sensor_data/meteo_10min.dat")
meteo_raw <- read.table(
  meteo_file,
  sep = ",",
  header = FALSE,
  skip = 4,
  stringsAsFactors = FALSE,
  col.names = c("datetime", "id", "air_temp", "rh", "patm", "vpd", "radiation")
)

meteo_daily <- meteo_raw %>%
  mutate(
    datetime = ymd_hms(datetime),
    date = as.Date(datetime)
  ) %>%
  mutate(across(c(air_temp, rh, vpd, radiation), as.numeric)) %>%
  group_by(date) %>%
  summarise(
    air_temp = mean(air_temp, na.rm = TRUE),
    rh = mean(rh, na.rm = TRUE),
    vpd = mean(vpd, na.rm = TRUE),
    radiation = mean(radiation, na.rm = TRUE),
    .groups = "drop"
  )

meteo_daily %>% slice_head(n = 5)
```

## SWP to daily site signal

```{r swp-daily}
soil_file <- .resolve_path("data/raw/sensor_data/soil_10min.dat")
soil_raw <- read.table(
  soil_file,
  sep = ",",
  header = TRUE,
  skip = 2,
  stringsAsFactors = FALSE
) %>%
  slice(-1)

soil_daily <- soil_raw %>%
  mutate(
    datetime = ymd_hms(TS),
    date = as.Date(datetime)
  ) %>%
  select(date, starts_with("kPa")) %>%
  mutate(across(-date, ~ as.numeric(na_if(.x, "NAN")))) %>%
  group_by(date) %>%
  summarise(across(everything(), \(x) mean(x, na.rm = TRUE)), .groups = "drop")

swp_site_daily <- soil_daily %>%
  pivot_longer(
    cols = starts_with("kPa"),
    names_to = "sensor_raw",
    values_to = "soil_mp"
  ) %>%
  filter(!is.na(soil_mp), soil_mp < 0) %>%
  mutate(swp_log10 = log10(-soil_mp)) %>%
  group_by(date) %>%
  summarise(
    swp_site = median(swp_log10, na.rm = TRUE),
    swp_site_sd = sd(swp_log10, na.rm = TRUE),
    .groups = "drop"
  )

swp_site_daily %>% slice_head(n = 5)
```

## SWC observations and modeling panel

```{r panel-build}
swc_obs <- readr::read_csv(
  .resolve_path("data/interim/box_soilwater.csv"),
  show_col_types = FALSE
) %>%
  transmute(
    boxlabel = as.character(boxlabel),
    date = as.Date(date),
    swc_obs = as.numeric(swc)
  )

date_min <- max(min(meteo_daily$date), min(swp_site_daily$date))
date_max <- min(max(meteo_daily$date), max(swp_site_daily$date))

all_boxes <- swc_obs %>% distinct(boxlabel)
all_dates <- tibble(date = seq.Date(date_min, date_max, by = "day"))

panel_daily <- tidyr::expand_grid(all_boxes, all_dates) %>%
  left_join(swc_obs, by = c("boxlabel", "date")) %>%
  left_join(meteo_daily, by = "date") %>%
  left_join(swp_site_daily %>% select(date, swp_site), by = "date") %>%
  mutate(
    date_num = as.numeric(date),
    boxlabel = factor(boxlabel)
  )

fit_data <- panel_daily %>%
  filter(!is.na(swc_obs), !is.na(swp_site), !is.na(air_temp), !is.na(vpd), !is.na(radiation))

coverage <- tibble::tribble(
  ~Metric, ~Value,
  "Date window", paste0(as.character(date_min), " to ", as.character(date_max)),
  "Total boxes", as.character(n_distinct(panel_daily$boxlabel)),
  "Observed SWC rows", as.character(sum(!is.na(panel_daily$swc_obs))),
  "Rows used for fitting", as.character(nrow(fit_data))
)

coverage
```

# Exploratory Figures

## Site-level SWP and climate

```{r site-signals}
p1 <- ggplot(swp_site_daily, aes(date, swp_site)) +
  geom_line(color = "firebrick", linewidth = 0.6) +
  labs(title = "Daily Site SWP Signal", x = "Date", y = "log10(-SWP)") +
  theme_linedraw()

p2 <- meteo_daily %>%
  pivot_longer(c(air_temp, rh, vpd, radiation), names_to = "var", values_to = "value") %>%
  ggplot(aes(date, value)) +
  geom_line(color = "steelblue", linewidth = 0.45) +
  facet_wrap(~ var, scales = "free_y", ncol = 2) +
  labs(title = "Daily Climate Covariates", x = "Date", y = NULL) +
  theme_linedraw()

p1
p2
```

## SWC sampling sparsity

```{r swc-sparsity}
swc_obs %>%
  filter(!is.na(swc_obs)) %>%
  ggplot(aes(date, fct_reorder(boxlabel, boxlabel))) +
  geom_point(alpha = 0.5, size = 0.6, color = "black") +
  labs(
    title = "Observed SWC Measurement Dates by Box",
    x = "Date",
    y = "Box"
  ) +
  theme_linedraw() +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

# Model Fitting

Agnostic GAM specification:

`swc_obs ~ s(date_num) + s(swp_site) + s(air_temp) + s(vpd) + s(radiation) + s(boxlabel, bs = "re")`

```{r fit-gam}
gam_swc <- mgcv::gam(
  swc_obs ~
    s(date_num, k = 10) +
    s(swp_site, k = 8) +
    s(air_temp, k = 8) +
    s(vpd, k = 8) +
    s(radiation, k = 8) +
    s(boxlabel, bs = "re"),
  data = fit_data,
  method = "REML"
)

summary(gam_swc)
```

## Smooth terms

```{r smooth-plots, fig.width=11, fig.height=8}
plot(gam_swc, pages = 1, shade = TRUE, residuals = TRUE, scale = 0)
```

# Prediction and Diagnostics

```{r predict}
pred_data <- panel_daily %>%
  filter(!is.na(swp_site), !is.na(air_temp), !is.na(vpd), !is.na(radiation))

pred <- predict(gam_swc, newdata = pred_data, se.fit = TRUE, type = "response")

out <- pred_data %>%
  mutate(
    swc_hat = as.numeric(pred$fit),
    swc_hat_se = as.numeric(pred$se.fit),
    swc_hat_lo = pmax(swc_hat - 1.96 * swc_hat_se, 0),
    swc_hat_hi = pmin(swc_hat + 1.96 * swc_hat_se, 100),
    swc_hat = pmin(pmax(swc_hat, 0), 100),
    swc_source = if_else(!is.na(swc_obs), "measured", "imputed_gam")
  ) %>%
  arrange(boxlabel, date)

metrics <- out %>%
  filter(!is.na(swc_obs)) %>%
  summarise(
    n = n(),
    rmse = sqrt(mean((swc_obs - swc_hat)^2)),
    mae = mean(abs(swc_obs - swc_hat)),
    r = cor(swc_obs, swc_hat)
  )

metrics
```

## Observed vs fitted

```{r obs-vs-fit}
out %>%
  filter(!is.na(swc_obs)) %>%
  ggplot(aes(swc_obs, swc_hat)) +
  geom_point(alpha = 0.35, size = 1.1, color = "grey25") +
  geom_abline(slope = 1, intercept = 0, color = "firebrick", linewidth = 0.7) +
  coord_equal() +
  labs(
    title = "Observed vs Fitted SWC (Measured Dates)",
    x = "Observed SWC",
    y = "Fitted SWC"
  ) +
  theme_linedraw()
```

## Residual checks

```{r residuals}
out_res <- out %>%
  filter(!is.na(swc_obs)) %>%
  mutate(resid = swc_obs - swc_hat)

p_res1 <- ggplot(out_res, aes(swc_hat, resid)) +
  geom_point(alpha = 0.35, size = 1.0) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "firebrick") +
  labs(title = "Residuals vs Fitted", x = "Fitted SWC", y = "Residual") +
  theme_linedraw()

p_res2 <- ggplot(out_res, aes(resid)) +
  geom_histogram(bins = 35, fill = "grey70", color = "white") +
  labs(title = "Residual Distribution", x = "Residual", y = "Count") +
  theme_linedraw()

p_res1
p_res2
```

## Example box trajectories

```{r example-trajectories, fig.width=11, fig.height=7}
set.seed(42)
example_boxes <- out %>% distinct(boxlabel) %>% slice_sample(n = 6) %>% pull(boxlabel)

out %>%
  filter(boxlabel %in% example_boxes) %>%
  ggplot(aes(date, swc_hat)) +
  geom_ribbon(aes(ymin = swc_hat_lo, ymax = swc_hat_hi), fill = "grey80", alpha = 0.5) +
  geom_line(color = "steelblue", linewidth = 0.5) +
  geom_point(
    data = ~ dplyr::filter(.x, !is.na(swc_obs)),
    aes(y = swc_obs),
    color = "black",
    size = 1.1,
    alpha = 0.8
  ) +
  facet_wrap(~ boxlabel, scales = "free_y", ncol = 3) +
  labs(
    title = "Example Daily SWC Trajectories (Lines = GAM, Points = Measured)",
    x = "Date",
    y = "SWC"
  ) +
  theme_linedraw()
```

# Export

```{r export}
out_file <- .resolve_path("data/interim/box_soilwater_daily_gam_agnostic.csv")
readr::write_csv(out, out_file)

tibble(
  output_file = out_file,
  n_rows = nrow(out),
  n_measured = sum(out$swc_source == "measured"),
  n_imputed = sum(out$swc_source == "imputed_gam")
)
```

# Session Info

```{r session-info}
sessionInfo()
```
