---
title: 'ALINV – Size Trajectories and Size Models'
date: "`r format(Sys.Date(), '%d %b %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
    highlight: espresso
    theme: journal
    fig_width: 9
    fig_height: 6
execute:
  warning: false
  message: false
  echo: true
  results: hide
params:
  scenario_label: "both soils (with soil as treatment)"
  soil_filter: "both"
  include_soil_treatment: true
  analysis_date: !r as.character(Sys.Date())
  output_root: "output"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  results = "hide",
  fig.show = "hold",
  fig.width = 9,
  fig.height = 6
)

source_path_candidates <- c(
  "functions/_source.R",
  "../functions/_source.R",
  "./functions/_source.R"
)
source_path <- source_path_candidates[file.exists(source_path_candidates)][1]
if (is.na(source_path)) {
  stop("Could not locate functions/_source.R for notebook setup.")
}
source(source_path)

source_dir <- dirname(source_path)
proj_root <- normalizePath(file.path(source_dir, ".."), winslash = "/", mustWork = TRUE)

source(file.path(source_dir, "1-summary-figures.R"))
source(file.path(source_dir, "3-effect-size-factorial.R"))
source(file.path(source_dir, "4-structural-equation-model.R"))
source(file.path(source_dir, "5-combining-glmm-sem.R"))
source(file.path(source_dir, "8-size-trajectories.R"))

analysis_ctx <- alinv_init_notebook_context(if (exists("params")) params else NULL)
scenario_label <- analysis_ctx$scenario_label
soil_type <- analysis_ctx$soil_filter
include_soil_treatment <- analysis_ctx$include_soil_treatment
analysis_root <- analysis_ctx$analysis_root
analysis_data_root <- analysis_ctx$analysis_data_root
species_vec <- c("fagus", "quercus")

options(alinv.temporal_y_limits = alinv_temporal_effect_y_limits())
```

# Introduction

This notebook focuses on the size-related responses in the ALINV experiment. It covers `Height`, `Diameter`, and `Volume`, where `Volume` is calculated as a cylindrical stem-volume proxy from the cleaned height and diameter measurements:

$$\mathrm{Volume\ (cm^3)} = \pi \left(\frac{\mathrm{Diameter\ (mm)}}{20}\right)^2 \times \mathrm{Height\ (cm)}$$

Current render scenario: `r scenario_label`. Soil filtering is set to `r soil_type`, and soil is `r if (include_soil_treatment) "included as an explicit treatment contrast." else "pooled across both soils and omitted as a treatment contrast."`

For the within-phase increments, phase 1 is referenced to the first measurement, while later phases are referenced to the last measurement of the previous phase. To make these resets visually explicit, the timeseries below add a zero-valued restart point at each phase boundary and break the line between phases.

Scenario-local plot-ready trajectory exports are written to `r alinv_project_relative_path(file.path(analysis_data_root, "size_trajectories"))`.

# Plot-Ready Data

```{r size-trajectory-exports}
size_bundle <- build_growth_plot_bundle(
  soil_type = soil_type,
  include_soil_treatment = include_soil_treatment
)
export_growth_plot_bundle(
  size_bundle,
  export_dir = alinv_data_path("size_trajectories", create_dir = TRUE)
)
```

# Size Responses

The model panels below pair the descriptive timeseries with the temporal GLMM and direct-effect SEM for each species. Positive effect sizes indicate beneficial effects and negative effect sizes indicate harmful effects.

```{r size-response-loop, results='asis', fig.width=9.5, fig.height=5.8}
size_specs <- growth_metric_catalog()

for (metric_name in unique(size_specs$metric_group)) {
  cat("## ", metric_name, "\n\n", sep = "")

  specs_metric <- size_specs %>%
    dplyr::filter(.data$metric_group == metric_name)

  for (i in seq_len(nrow(specs_metric))) {
    spec_i <- specs_metric[i, ]

    cat("### ", spec_i$variant_label[[1]], "\n\n", sep = "")
    print(
      plot_growth_metric_ts(
        resp_var = spec_i$resp_var[[1]],
        y_label = spec_i$y_label[[1]],
        panel_title = spec_i$panel_title[[1]],
        soil_type = soil_type,
        include_soil_treatment = include_soil_treatment,
        drought_bars = TRUE
      )
    )
    cat("\n")

    for (species_i in species_vec) {
      cat("#### ", stringr::str_to_title(species_i), "\n\n", sep = "")

      res_i <- make_temporal_sem_combo(
        type = "tree",
        data_name = "growth",
        resp_var = spec_i$resp_var[[1]],
        species = species_i,
        soil_type = soil_type,
        include_soil_treatment = include_soil_treatment,
        include_interaction = FALSE
      )

      print(res_i$temporal$plot)
      print(res_i$sem$plot)
      cat("\n")
    }
  }
}
```

# Session Info

```{r session-info}
sessionInfo()
```
