---
title: 'ALINV – SWP Provenance'
date: "`r format(Sys.Date(), '%d %b %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    highlight: espresso
    theme: journal
    fig_width: 9
    fig_height: 6
execute:
  warning: false
  message: false
  echo: true
  results: hide
params:
  scenario_label: "global"
  soil_filter: "both"
  include_soil_treatment: false
  analysis_date: !r as.character(Sys.Date())
  output_root: "output"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  results = "hide",
  fig.show = "hold",
  fig.width = 9,
  fig.height = 6
)

source_path_candidates <- c(
  "functions/_source.R",
  "../functions/_source.R",
  "./functions/_source.R"
)
source_path <- source_path_candidates[file.exists(source_path_candidates)][1]
if (is.na(source_path)) {
  stop("Could not locate functions/_source.R for notebook setup.")
}
source(source_path)

source_dir <- dirname(source_path)
proj_root <- normalizePath(file.path(source_dir, ".."), winslash = "/", mustWork = TRUE)

source(file.path(source_dir, "1-summary-figures.R"))

analysis_ctx <- alinv_init_notebook_context(if (exists("params")) params else NULL)
analysis_root <- analysis_ctx$analysis_root
analysis_data_root <- analysis_ctx$analysis_data_root
swp_export_dir <- alinv_data_path("swp_provenance", create_dir = TRUE)
```

# Introduction

This notebook documents how the final soil water potential (SWP) product is derived from the raw sensor timeseries and how it becomes the final figures used in the summary notebooks.

All intermediate tables created here are exported under `r alinv_project_relative_path(swp_export_dir)`.

# 1. Raw Sensor Timeseries to Daily Long Table

The raw input is the 10-minute sensor file `data/raw/sensor_data/soil_10min.dat`. The wrangling helper:

1. reads the raw file,
2. aggregates it to daily mean SWP per sensor,
3. maps sensors to mesocosm boxes,
4. joins the box metadata needed for treatment-level summaries.

```{r swp-daily-long}
soil_daily_long <- wrangle_soil_data()
readr::write_csv(soil_daily_long, file.path(swp_export_dir, "soil-water-potential_daily-long.csv"))
```

```{r swp-daily-long-summary}
soil_daily_summary <- soil_daily_long %>%
  dplyr::summarise(
    n_dates = dplyr::n_distinct(.data$date),
    n_boxes = dplyr::n_distinct(.data$boxlabel),
    n_rows = dplyr::n(),
    n_missing_swp = sum(is.na(.data$soil_mp))
  )

knitr::kable(soil_daily_summary, caption = "Daily long-table summary after sensor aggregation and metadata joins")
```

# 2. Filter and Transform to the Plot Scale

Only negative SWP values are retained, and the final plotted scale is:

$$\mathrm{SWP}_{plot} = -\log_{10}(-\mathrm{SWP}_{kPa})$$

This is the scale used by the summary notebooks for the SWP figures.

```{r swp-plot-scale}
soil_plot_scale <- soil_daily_long %>%
  dplyr::filter(!is.na(.data$soil_mp), .data$soil_mp < 0) %>%
  dplyr::mutate(
    soil_mp_plot = -log10(-.data$soil_mp),
    soiltype = gsub("-", "_", .data$soiltype)
  )

readr::write_csv(soil_plot_scale, file.path(swp_export_dir, "soil-water-potential_plot-scale.csv"))
```

```{r swp-plot-scale-summary}
plot_scale_summary <- soil_plot_scale %>%
  dplyr::summarise(
    n_rows = dplyr::n(),
    min_plot_value = min(.data$soil_mp_plot, na.rm = TRUE),
    max_plot_value = max(.data$soil_mp_plot, na.rm = TRUE)
  )

knitr::kable(plot_scale_summary, caption = "Summary of the filtered SWP values on the final plot scale")
```

# 3. Daily Treatment Summaries Feeding the Final Figures

The summary figures in the main notebooks then aggregate the plot-scale SWP by date and treatment combination. These are the plot-ready daily means used downstream for the raw and smoothed trajectory figures.

```{r swp-treatment-summary}
soil_plot_daily_grouped <- soil_plot_scale %>%
  dplyr::group_by(
    .data$date,
    .data$soiltype,
    .data$robinia,
    .data$precipitation,
    .data$culture
  ) %>%
  dplyr::summarise(
    soil_mp_plot_mean = mean(.data$soil_mp_plot, na.rm = TRUE),
    n_boxes = dplyr::n_distinct(.data$boxlabel),
    .groups = "drop"
  )

readr::write_csv(soil_plot_daily_grouped, file.path(swp_export_dir, "soil-water-potential_plot-daily-treatment-means.csv"))
```

```{r swp-treatment-summary-preview}
knitr::kable(
  dplyr::slice_head(soil_plot_daily_grouped, n = 12),
  caption = "Preview of the daily treatment-level SWP table used for the downstream plots"
)
```

# 4. Final Figures Used Downstream

These are the same final SWP figures that are called in the summary notebooks.

## Raw Trajectories Plus Smoothed Treatment Means

```{r swp-final-both, fig.width=10, fig.height=6}
plot_soil_mp(
  soil_daily_long,
  filter_soiltype = "both",
  show = "both",
  include_soil_treatment = TRUE
)
```

## Raw Treatment Means

```{r swp-final-raw, fig.width=10, fig.height=6}
plot_soil_mp(
  soil_daily_long,
  filter_soiltype = "both",
  show = "raw",
  include_soil_treatment = TRUE
)
```

# Session Info

```{r session-info}
sessionInfo()
```
